% !TEX root = thesis.tex
% =============================================================================
% Chapter 3 — Related Work
% Zero-Shot Nutrient and Quantity Association using OCR-Guided Semantic Graph Matching
% Moustafa Hamada, THD + USB
%
% Required packages (already loaded by thesis.tex):
%   inputenc (utf8), graphicx, amsmath/amssymb, booktabs, url,
%   hyperref, libertine.
%
% Cross-references to Chapter 2 use the sec:bg-* label scheme.
%
% Chapter-2 labels referenced from this chapter:
%   sec:bg-problem                 (= §2.1)
%   sec:bg-ocr-entry               (= §2.2.1)
%   sec:bg-class-rules             (= §2.4.2)
%   sec:bg-class-zeroshot          (= §2.4.5)
%   sec:bg-graphs                  (= §2.5)
%   sec:bg-graphs-intro            (= §2.5.1)
%   sec:bg-graphs-detlearn         (= §2.5.5)
%   sec:bg-embeddings              (= §2.6)
%   sec:bg-evaluation-twotier      (= §2.7.4)
%   sec:bg-evaluation-diagnostics  (= §2.7.5)
% =============================================================================

\chapter{Related Work}
\label{ch:related-work}

This chapter does the work Chapter~2 deferred. Chapter~2 introduced the technical primitives that the pipeline of this thesis depends on, restricting itself to the conceptual machinery the methodology and experimental chapters require, and leaving to the present chapter the question of how the thesis's design choices relate to the published systems that share its problem space. Chapter~3 takes up that deferred question. It surveys the research traditions the thesis sits within --- visually rich document understanding, key information extraction from receipts and forms, zero-shot named-entity recognition, the application domain of nutrition-label extraction, and the methodologies for evaluating systems of this kind --- and identifies the specific point in the design space the thesis occupies. Where a system has already been introduced in Chapter~2 --- PICK, SPADE, GLiNER, BGE-M3 --- the present discussion extends rather than duplicates the earlier treatment, focusing on the system-level positioning rather than the architectures Chapter~2 has already explained. Where a system is introduced for the first time --- the LayoutLM family, Donut, Pix2Struct, the form-and-receipt benchmark lineage, the LLM-as-judge evaluation literature --- the chapter supplies enough detail to make the comparison legible without restating common knowledge a reader of the document-understanding literature would already possess.

The chapter is organised as follows. \S\ref{sec:supervised_doc_ie} surveys the broader landscape of visually rich document understanding, contrasting the \textbf{cascaded-pipeline tradition} this thesis belongs to against the \textbf{end-to-end visual document understanding} paradigm that has come to dominate the document-AI literature, and explains why a pipeline approach remains the right choice for a labelled-data-scarce target domain. \S\ref{sec:tsr} narrows to table structure recognition specifically, treating it as an alternative paradigm whose extraction target differs from the thesis's. \S\ref{sec:vlm_ocr_free} covers OCR-free and vision-language document understanding as a second alternative paradigm that removes the staging the thesis retains. \S\ref{sec:graph_kie} returns to the graph-based key-information-extraction lineage that Chapter~2 borrowed from at the primitive level, and articulates the system-level positioning. \S\ref{sec:zero_shot_ie} covers the zero-shot information-extraction research line on its own terms, distinguishing schema-flexible NER models from embedding-prototype matching and from LLM-prompted extraction. \S\ref{sec:nutrition_label_extraction} discusses the application domain --- nutrition labels and adjacent food-data extraction --- and is candid about the relative scarcity of prior work on this specific target. \S\ref{sec:thesis_positioning} closes by stating where the thesis sits within the surveyed landscape.

% =============================================================================
\section{Supervised document information extraction and benchmark-driven progress}
\label{sec:supervised_doc_ie}

\subsection{The document-IE problem and its benchmark substrate}
\label{sec:doc_ie_problem}

Document information extraction is the task of converting a document image, or a digital document with preserved layout, into a structured record of the facts the document contains. Unlike plain-text information extraction, the input carries spatial structure that is part of the meaning: a value placed inside a box on a form is the answer to the question printed beside the box; a number positioned in a table cell is the value of the column header above it and the row label to its left; an entry separated from the next by horizontal whitespace belongs to a different logical field. A document-IE system therefore consumes three signals --- the textual content of the document, the geometry of where each piece of text sits, and, in many systems, the visual appearance of the document image itself --- and produces some target structured representation: a tagged sequence of tokens, a set of typed entity spans, a list of key--value pairs, or a fully populated record schema.

The progress of the field over the last decade has been driven by a small number of benchmark datasets, each of which fixes a specific instantiation of the problem. \textbf{FUNSD}~\cite{jaume2019funsd} is a canonical benchmark for form understanding: 199 noisy scanned forms with token-level annotations of four entity classes (header, question, answer, other) and explicit linking annotations between question and answer pairs. The benchmark rewards two related tasks --- entity tagging and entity linking --- and has become a standard test bed for the form-understanding subset of document IE. \textbf{SROIE}~\cite{huang2019sroie} is the scanned-receipt benchmark introduced at ICDAR 2019, with three tasks defined on the same underlying images: text localisation, OCR transcription, and key-information extraction over a four-field schema (company name, date, address, total). \textbf{CORD}~\cite{park2019cord} is a large receipt parsing dataset of Indonesian receipts introduced shortly afterwards, with multi-level semantic annotations: 30 fine-grained field labels organised into 8 super-classes, together with OCR boxes and reading-order annotations. Together these three datasets --- forms, receipts at the field level, receipts at the parsing level --- constitute a commonly used benchmark cluster for document KIE and receipt-and-form understanding, and the systems that have come to define the modern document-IE literature are in large part the systems that have done well on them.

A fourth dataset broadens the benchmark lineage in a different direction. \textbf{DocBank}~\cite{li2020docbank} is a large-scale token-level annotated corpus of approximately five hundred thousand document pages drawn from arXiv submissions, with fine-grained layout annotations covering twelve component classes such as title, paragraph, list, table, figure, and equation. DocBank is not a KIE benchmark in the FUNSD/SROIE/CORD sense --- it does not annotate typed key--value pairs or a target record schema --- but a layout-analysis benchmark that asks systems to identify the structural role of each token in a document. Including it here is methodologically informative rather than directly comparative: while FUNSD, SROIE, and CORD represent KIE-oriented benchmarks, DocBank broadens the benchmark lineage toward token-level document layout analysis, and the systems that have done well on it tend to share architectural commitments with KIE systems in ways the next subsection draws out.

These benchmarks have made forms, receipts, and scientific documents central to document-IE evaluation, but nutrition and supplement labels are not part of the canonical benchmark set, and \S\ref{sec:nutrition_label_extraction} returns to that absence and what it has meant for the small body of nutrition-specific extraction work that exists outside the benchmark culture.

\subsection{The supervised paradigm and what it assumes}
\label{sec:supervised_paradigm}

What unites the systems that have come to define benchmark-driven progress in document IE is not a single architecture but a shared methodological regime. A model is first trained on large unlabelled document corpora using self-supervised objectives that exploit the document's text, layout, and often image content; the resulting parameters are then adapted to a labelled downstream dataset whose annotation schema defines the target output. The downstream dataset supplies both the \textit{task} --- entity tagging, key-value linking, full record extraction --- and the \textit{schema} against which performance is measured. This is the familiar pretraining-and-fine-tuning pattern, but in the document-IE setting it is made multimodal by the addition of layout coordinates and image evidence to the textual input that natural-language pretraining typically operates on. The regime's hallmark is that the schema the system is evaluated against is the schema for which labelled training examples have been provided, and the model's downstream performance depends on the availability and quality of those examples.

A central model family within this regime is the \textbf{LayoutLM lineage}~\cite{xu2020layoutlm,xu2021layoutlmv2,huang2022layoutlmv3} together with several closely related transformer-based document-understanding architectures, including \textbf{DocFormer}~\cite{appalaraju2021docformer}, \textbf{TILT}~\cite{powalski2021tilt}, and \textbf{BROS}~\cite{hong2022bros}. Most of these systems combine textual, layout, and visual evidence into a single transformer encoder pretrained on large-scale unlabelled document corpora; others, such as BROS, show that strong KIE performance can also be obtained by concentrating on text and two-dimensional layout alone, with carefully designed spatial encodings standing in for the image-stream signal the multimodal models exploit. A representative example is the masked visual-language modelling objective used in the LayoutLM family and extended in LayoutLMv2~\cite{xu2021layoutlmv2}, where selected text tokens are masked and the model must recover them from surrounding textual, spatial, and visual context; LayoutLMv2 further adds text--image alignment and text--image matching objectives to strengthen cross-modal interaction during pretraining. After pretraining, the same encoder is fine-tuned on FUNSD-style entity-linking, SROIE-style or CORD-style KIE, or other downstream targets, with task-specific heads attached to the shared backbone.

The strength of this regime is also its condition. The benefits are real: end-to-end optimisation against a benchmark-defined objective tends to produce systems that are well-tuned to that objective; multimodal pretraining permits joint reasoning over modalities that would be hard to integrate by composing separate modules; and the leaderboard culture surrounding the benchmarks of \S\ref{sec:doc_ie_problem} has produced a steady accumulation of architectural and training-recipe improvements that compound over successive model generations. The condition is that the regime presupposes labelled examples in the target schema. Adapting one of these systems to a new schema for which no labelled corpus exists requires either annotating one --- which the authors of FUNSD, SROIE, CORD, and DocBank did at substantial cost --- or some indirect path through cross-task transfer or zero-shot evaluation that the regime as such does not natively provide. The regime also depends on schema stability between annotation, fine-tuning, and deployment: if the target schema changes, labelled data and task heads often have to be revised. These are conditions under which the regime's strengths are realised, and they are what \S\ref{sec:thesis_position_supervised_regime} takes up next.

\subsection{Where this thesis sits relative to the supervised regime}
\label{sec:thesis_position_supervised_regime}

The regime described in \S\ref{sec:supervised_paradigm} presupposes labelled examples in the target schema. This thesis works on a domain --- EU supplement nutrition labels --- for which no canonical labelled benchmark in the FUNSD, SROIE, CORD, or DocBank lineage exists, and is committed to operating zero-shot apart from the limited validation-based threshold tuning described in \S\ref{sec:bg-problem}. The consequence is direct: applying the supervised regime to this thesis's setting would first require constructing the labelled corpus the thesis is committed to not presupposing, which would contradict the zero-shot commitment that motivates the architectural choices the rest of the thesis describes. The work of the remaining sections is therefore to examine research lines that either relax the labelled-data requirement, change the extraction target, or offer representations that can be re-implemented without supervised training.

The neighbouring research lines fall into three structural responses to this mismatch. The first is to \textit{change the extraction target} by routing the problem through an architectural paradigm whose output is not the same record schema that supervised KIE consumes: \textbf{table structure recognition} (\S\ref{sec:tsr}) reconstructs the cell grid of a tabular region as an intermediate representation from which records can be derived downstream, and \textbf{OCR-free vision-language document understanding} (\S\ref{sec:vlm_ocr_free}) folds OCR and IE together into a single end-to-end model that emits the structured output directly from the document image. The second is to \textit{retain the representation but re-implement it without supervised training}: \textbf{graph-based key information extraction} (\S\ref{sec:graph_kie}), which is the lineage \S\ref{sec:bg-graphs} has already drawn from, supplies the typed-edge document representation that the thesis's association stage uses, with the deterministic-versus-supervised choice \S\ref{sec:bg-graphs-detlearn} named at the level of individual primitives now revisited at the level of complete systems. The third is to \textit{operate without labelled target-schema data in the first place}: \textbf{zero-shot information extraction and semantic matching} (\S\ref{sec:zero_shot_ie}) covers the schema-flexible NER lineage, prototype-based matching, and LLM-prompted extraction, all of which produce structured output by means other than fitting parameters to a labelled training set. The domain-specific check that completes the survey --- whether the supplement-label setting has been previously addressed at all, and how the small body of nutrition-label-extraction work that exists relates to the canonical benchmark culture --- is the work of \S\ref{sec:nutrition_label_extraction}.

The constraint set that defines what this thesis needs from these research lines is the one introduced in \S\ref{sec:bg-problem}: the system should operate without target-schema training data, remain flexible across the semi-structured layouts found in supplement labels, and preserve enough modularity for errors to be inspected at the OCR, correction, classification, and association stages. The rest of this chapter therefore examines how neighbouring research lines --- table-structure recognition, OCR-free document VLMs, graph-based KIE, zero-shot extraction, and nutrition-label parsing --- meet, evade, or leave unresolved the constraints that define the supplement-label setting.

% =============================================================================
\section{Table structure recognition}
\label{sec:tsr}

\subsection{The table structure recognition problem and its representations}
\label{sec:tsr_problem}

Table structure recognition (TSR) is the task of taking a region known or believed to contain a table and producing a structured representation of the table's grid. The input is the table image, optionally accompanied by OCR-recognised text and bounding boxes; the output is a description of the table's structural composition, including the positions and adjacency of cells, their membership in rows and columns, the relations between header and data cells, and any cell spans that cross the basic row--column grid. TSR is distinct from the two problems Chapter~2 has already covered. OCR maps pixels to text; KIE maps text and layout to typed records; TSR maps a table region to a grid representation that sits between the two and commits to neither typed semantics nor raw transcription as its target output.

The representations TSR systems emit fall into a small number of families. HTML-like markup is one widely used target representation, in part because PubTabNet~\cite{zhong2019pubtabnet} formulates image-based table recognition as the conversion of table images into corresponding HTML structure expressed through \texttt{\textless tr\textgreater}, \texttt{\textless td\textgreater}, and \texttt{\textless th\textgreater} tags with \texttt{rowspan} and \texttt{colspan} attributes for cell spans. A second representation is the logical--physical structure split, in which logical structure --- which cell belongs to which row and column, which cells are headers, where spans occur --- is separated from physical structure, the bounding-box coordinates of each cell in the image. TSR systems differ in whether they emit logical structure, physical structure, or both, and the two are sometimes produced by separate sub-modules within a single system. What unites these representations is what they describe: they describe the grid of the table --- its cells, its rows, its columns, its headers, its spans --- not the domain meaning of the cells beyond marking some of them as headers. The cells contain text after OCR has been applied, but TSR commits to no further typing of that text into application-level categories. This distinction is what the positioning argument in \S\ref{sec:why_no_tsr} turns on.

TSR is evaluated by metrics that reflect the same grid-versus-meaning distinction. TEDS~\cite{zhong2019pubtabnet} --- Tree-Edit-Distance-based Similarity --- evaluates the similarity between predicted and reference table structures through tree-edit distance over HTML-like table representations, computing how many insertions, deletions, and substitutions of nodes are required to transform the predicted markup into the reference markup. GriTS~\cite{smock2023grits} --- Grid Table Similarity --- evaluates a predicted table more directly as a grid or matrix of cells, comparing predicted and reference structures in the natural matrix representation rather than only through markup edit distance, and can be applied to cell topology, cell location, and cell content recognition within the same grid-based framework. Both metrics are TSR-internal: they evaluate the recovered grid, not whether the cells contain the right typed records under any application-level schema.

\subsection{Representative TSR systems and their architectural commitments}
\label{sec:tsr_systems}

The architectural surface of TSR can be organised along the axis of how a system produces its grid representation. \textit{Separator-and-boundary-based} TSR predicts row separators, column separators, or cell boundaries on the image, and assembles cells from the predicted lines or boxes. \textit{Sequence-generation-based} TSR emits the table structure as a structured token sequence --- typically as HTML or as an HTML-like markup --- produced by a transformer decoder over the table-image features. A third family, \textit{hybrid systems}, predicts logical structure and physical cell locations jointly, often through dual decoding paths that share an encoder. The choice between these families has consequences for how the system handles spanning cells, irregular grids, and dense-text content, and the canonical TSR systems sit at different points on this axis depending on the trade-offs their authors prioritised.

Two systems are representative for the purposes of this section. \textbf{TableFormer}~\cite{nassar2022tableformer} uses a transformer-based encoder--decoder design in which one decoding path predicts the table structure as an HTML-like token sequence while a cell-bounding-box decoder predicts the physical location of table cells, jointly targeting logical structure and physical cell localisation in a single supervised end-to-end model. \textbf{SLANet} represents the lightweight production-oriented line of TSR systems in the PaddleOCR ecosystem; it predicts table structure and cell locations through a structure--location-alignment design rather than through separator-based assembly, with a PP-LCNet backbone, CSP-PAN feature fusion, and an SLA head that aligns structure and position information. SLANet is one of several modules integrated into PP-StructureV3~\cite{cui2025paddleocr3}, the hierarchical document-parsing pipeline distributed alongside PP-OCRv5 in the PaddleOCR 3.0 release, with later SLANeXt variants providing dedicated weights for wired and wireless table recognition~\cite{ppstructurev3}. This point is especially relevant because the OCR engine used in this thesis comes from the PaddleOCR ecosystem, whose document-parsing stack includes PP-StructureV3 and the table-recognition modules just described. The thesis therefore does not ignore TSR because it was unavailable within the chosen ecosystem; it omits it because invoking TSR would add an explicit grid-reconstruction stage to the pipeline, and \S\ref{sec:why_no_tsr} sets out the specific reasons that addition was not made.

The benchmark substrate commonly used for TSR training and evaluation includes three datasets that are especially relevant here. \textbf{PubTabNet}~\cite{zhong2019pubtabnet} contains approximately 568,000 table images extracted from scientific articles, annotated with corresponding HTML structure, and has become one of the central benchmarks for image-based table recognition. \textbf{FinTabNet}~\cite{zheng2021gte} provides cell-level annotations for tables drawn from financial documents. \textbf{TableBank}~\cite{li2020tablebank} is a weakly supervised benchmark built by extracting tables from Word and LaTeX documents, containing approximately 417,000 labelled tables, with structure derived from the source markup rather than from human annotation. These three benchmarks have shaped the evaluation culture of TSR over the last several years and define what ``good TSR performance'' means in the published literature. TSR systems trained and evaluated on these benchmarks are therefore optimised for table distributions shaped by scientific, financial, or typeset document conventions, rather than by the photographed, curved, multilingual, and semi-tabular layouts that supplement labels often exhibit.

\subsection{Why this thesis does not include a TSR stage}
\label{sec:why_no_tsr}

The exclusion of a dedicated TSR stage is a deliberate architectural choice rather than an omission. Three reasons motivate it: TSR's output schema does not match the thesis's target records; state-of-the-art TSR systems depend on supervised table-structure training data drawn from distributions unlike supplement labels; and a TSR-first pipeline would complicate layout handling by requiring routing between tabular and non-tabular cases.

The first reason is schema mismatch. As \S\ref{sec:tsr_problem} established, TSR produces a grid representation: cells, rows, columns, headers, spans. The thesis produces typed records of the form $\langle$nutrient, quantity, unit, context$\rangle$, in which each field is assigned an application-level semantic role. Even a perfect TSR system would leave the typing problem --- which cell is the nutrient name, which is the quantity --- to a downstream stage that TSR systems are not themselves evaluated on. A TSR-first pipeline could still be built, but it would be a two-stage system: grid reconstruction followed by semantic interpretation of the recovered rows, columns, and headers. The thesis chose to do that semantic interpretation directly on the OCR token stream by means of the geometric and graph-based machinery introduced in Chapter~2, rather than materialise an intermediate grid representation that would still need to be typed before the target records emerged.

The second reason is supervision dependency. The TSR systems surveyed in \S\ref{sec:tsr_systems} are supervised: they are trained on PubTabNet, FinTabNet, TableBank, and similar benchmarks, and their performance depends on the availability and distributional alignment of labelled table-structure training data. The closing observation of \S\ref{sec:tsr_systems} already established that those benchmarks represent table distributions shaped by scientific, financial, or typeset document conventions, while supplement labels are photographed, often curved, multilingual, and frequently semi-tabular. Evaluating the domain transfer of supervised TSR systems to supplement-label distributions would itself involve substantial empirical work outside the scope of this thesis, and using supervised TSR would in any case reintroduce the labelled-data dependency that the zero-shot commitment of \S\ref{sec:bg-problem} is designed to avoid. The thesis's design constraint is therefore to find primitives that operate without supervised target-domain data, which the geometric and graph-based reasoning of Chapter~2 supplies and which the supervised TSR systems of \S\ref{sec:tsr_systems} do not.

The third reason is layout-flexibility scope. Many supplement labels in the dataset are not arranged as clean fully-bordered grids: the corpus includes paragraph-prose nutrient declarations, single-column layouts, and multilingual slash-rows with values flowing alongside their multilingual nutrient names. A TSR-first pipeline would therefore need a layout-routing stage that decides whether a given label should be sent through TSR or handled by a non-tabular fallback path, and the fallback path would still need geometric reasoning over the OCR token stream of the kind \S\ref{sec:bg-graphs} describes. The thesis chose to treat tabular and non-tabular cases under a single unified geometric-graph framework rather than route them through different architectural paths. In the terms of \S\ref{sec:thesis_position_supervised_regime}, TSR is therefore an alternative that \textit{changes the extraction target}: it reconstructs a table before information extraction can begin. This thesis instead keeps the extraction target fixed as nutrient--quantity--unit--context association and treats tabular and non-tabular label layouts under a single geometric-graph representation.

% =============================================================================
\section{OCR-free and vision-language document understanding}
\label{sec:vlm_ocr_free}

\subsection{From OCR-dependent pipelines to OCR-free document models}
\label{sec:ocr_free_shift}

The default architectural pattern for visually rich document IE has for some years been a cascade of three stages. An optical character recogniser produces text together with bounding boxes from the document image; a layout-aware encoder consumes the OCR output together with geometric features and, in many systems, visual features extracted from the same image; and a task-specific head emits the target structured output. The cascade is what allows the textual, spatial, and visual signals introduced in \S\ref{sec:doc_ie_problem} to be brought together for downstream reasoning, and it is the architectural pattern within which the supervised regime of \S\ref{sec:supervised_paradigm} has accumulated its successive model generations. This is also the architectural family to which the pipeline developed in Chapter~2 belongs: OCR produces the token stream, post-OCR correction repairs it, and the later stages reason over the corrected tokens and their geometry rather than over the original image directly.

The cascaded pattern incurs three costs. The first, already identified in \S\ref{sec:bg-ocr-entry}, is \textbf{OCR error propagation}: errors at the perceptual stage are unrecoverable by any downstream component, and a misread character or a missing token simply does not appear in the output. The second is \textbf{staging overhead}. The cascade requires a separate OCR engine, a post-OCR correction stage to repair its characteristic error distribution, and an interface between the OCR output schema and whatever the next stage expects, each of which adds engineering surface area and creates an interface at which assumptions between modules must be made explicit. A further cost is \textbf{information loss at the OCR boundary}: once the image has been converted into recognised text and bounding boxes, downstream stages no longer have direct access to visual cues such as font weight, colour, texture, or local image evidence that might inform classification or association decisions. These costs were accepted as the necessary price of OCR-based pipelines until multimodal pretraining at scale produced encoders that could attend to image patches directly without going through a recognised-text intermediate.

The OCR-free alternative reverses the architectural commitment. Rather than recover text from the image and reason over the recovered text, an OCR-free document model consumes the image directly through a vision encoder and emits the structured target --- a tag sequence, a JSON-like record, a markup string --- through an autoregressive decoder, without committing to a separate OCR stage at all. The shift is enabled by three convergent developments: image-to-text encoder--decoder architectures from the broader vision-language literature, large-scale multimodal pretraining on document and screenshot data, and structured-output decoding objectives that bypass the intermediate text representation. The point is not that OCR-free models contain no text-like internal representations; it is that the pipeline no longer materialises OCR text as a separate, intermediate artefact that downstream components consume. The named systems that instantiate these design choices are the subject of \S\ref{sec:vlm_systems}.

\subsection{Representative OCR-free and vision-language systems}
\label{sec:vlm_systems}

The canonical OCR-free document understanding system is \textbf{Donut}~\cite{kim2022donut}, one of the clearest early examples of the paradigm in practice. Donut combines a Swin-Transformer image encoder over the document image with a BART-style autoregressive decoder that emits the structured target as a sequence of tokens, including special tags marking field boundaries within the output. The model has no OCR sub-module: the encoder consumes image patches directly and the decoder generates structured text without any intermediate recognised-text representation being materialised. The pretraining recipe uses synthetically generated document images and their token sequences, followed by task-specific fine-tuning on the downstream IE schema. Donut's authors frame the model explicitly as a response to the costs \S\ref{sec:ocr_free_shift} named: OCR cost, OCR inflexibility across languages and document types, and OCR error propagation. The system's contribution is to demonstrate that document IE can be performed end-to-end without an explicit OCR stage, given sufficient pretraining and supervised adaptation to the target schema.

A second representative system, \textbf{Pix2Struct}~\cite{lee2023pix2struct}, extends the image-to-text view beyond document forms by treating visually situated language understanding as a general image-to-sequence problem. Pix2Struct combines a vision-transformer encoder with a text decoder, but its distinguishing commitment is its pretraining objective: it learns to parse masked web screenshots into simplified HTML, training the model to map visual layout to structured text on a corpus drawn from the open web rather than from document-specific annotation. The pretraining task --- predicting the missing portions of a page's HTML from a partially masked screenshot --- produces a model whose representations are aligned with the structural conventions HTML encodes, and whose downstream behaviour on document IE, web UI understanding, infographic question answering, and related tasks is driven by that alignment rather than by document-specific inductive biases. Pix2Struct represents a different point in the OCR-free design space than Donut: the architectural shape is similar, but the pretraining commits the model to a more general image-to-structured-text capacity.

\textbf{LayoutLMv3}~\cite{huang2022layoutlmv3} is relevant here as a contrast rather than as an OCR-free system. It remains OCR-dependent, because it consumes recognised words together with layout coordinates and image patches as input, but it is mentioned in this section as a bridge: its unified text--image masking and word--patch alignment pretraining objectives show how document AI moved from text-and-layout encoders toward more tightly coupled visual-language pretraining, and the architectural commitments it foregrounded --- image patches as first-class input, cross-modal alignment as a pretraining objective --- are part of what made the OCR-free transition more natural. LayoutLMv3 itself is in the supervised regime of \S\ref{sec:supervised_paradigm} and consumes OCR output, but the design choices it made are visible in the OCR-free systems that followed.

A more recent use pattern is to prompt general-purpose document-capable vision-language models --- such as the \textbf{Qwen-VL}~\cite{bai2023qwenvl} and \textbf{InternVL}~\cite{chen2024internvl} families, or proprietary systems such as Gemini and GPT-4V/GPT-4o-style models --- to read a document image and emit JSON-like records in response to a natural-language instruction. In that use pattern, the model is not a document-specific OCR-free architecture in the Donut sense; the end-to-end extraction behaviour emerges from large-scale multimodal pretraining and instruction following, and the same model that handles a supplement label can be redirected to a receipt, a form, or a screenshot by changing the prompt. This is not the same architectural claim as Donut's OCR-free document transformer: it is an inference-time use pattern built on general multimodal instruction following. The distinction between specialised OCR-free document models and frontier VLMs used as extractors is what \S\ref{sec:why_explicit_ocr} sharpens into two separate positioning arguments.

\subsection{Why this thesis retains explicit OCR and modular extraction}
\label{sec:why_explicit_ocr}

The choice not to adopt either of the alternatives surveyed in \S\ref{sec:vlm_systems} --- specialised OCR-free document models or frontier vision-language models used as extractors --- is a deliberate architectural decision rather than an omission. Three reasons motivate it: schema-specific reliability and zero-shot control, modularity and interpretability, and deployment and reproducibility. The reasons split unevenly across the two \S\ref{sec:vlm_systems} categories. Schema-specific reliability is primarily a concern about specialised OCR-free models such as Donut and Pix2Struct, which require task-specific adaptation to produce reliable schema-conformant output. Deployment and reproducibility are primarily concerns about the frontier-VLM-as-extractor use pattern, which carries version, cost, privacy, and auditability considerations the deterministic pipeline avoids. Modularity and interpretability apply to both categories, because the end-to-end generation that defines either alternative hides the intermediate decisions a modular pipeline exposes.

The first reason concerns schema-specific reliability and zero-shot control. For specialised OCR-free models, the problem is that the systems surveyed in \S\ref{sec:vlm_systems} are pretrained on synthetic documents or web screenshots and require task-specific fine-tuning to produce reliable output in the target schema, which would reintroduce the labelled-data dependency that the zero-shot commitment of \S\ref{sec:bg-problem} is designed to avoid. The thesis would have to construct a labelled supplement-label corpus to fine-tune Donut or Pix2Struct against, which is precisely what the zero-shot commitment is designed to avoid. For frontier VLMs used as extractors, the situation is different: zero-shot prompting is available, but the resulting extraction behaviour exhibits its own control concerns, including variation in schema adherence across runs, model versions, and prompt formulations. The thesis's deterministic pipeline produces identical output for identical input by construction, which is a form of control that prompt-based extraction does not provide and that would itself become part of any maintenance regime built around it.

The second reason concerns modularity and interpretability, the third commitment of \S\ref{sec:bg-problem}. End-to-end models --- whether specialised OCR-free systems or frontier VLMs prompted at inference --- emit a structured output as a single autoregressive generation, and the perceptual, correction, classification, and association decisions that produced any given output are not separately observable. The four diagnostic CSVs \S\ref{sec:bg-evaluation-diagnostics} names --- \texttt{evaluation\_summary.csv}, \texttt{per\_image\_results.csv}, \texttt{tuples.csv}, and \texttt{pair\_details.csv} --- and the stage-level audit trail they constitute would have no direct analogue in an end-to-end system. A failure on a particular tuple cannot be traced to a specific OCR token, a specific correction rule, a specific classification decision, or a specific graph-association edge. The thesis chose modularity because that traceability is part of what makes single-variable ablations a viable methodology and what makes the experimental chapter's stage-level causal claims defensible.

The third reason concerns deployment and reproducibility, primarily for the frontier-VLM-as-extractor use pattern. Frontier vision-language models are deployed as services whose model versions are updated by their providers, so a result obtained against one snapshot may not reproduce against the next; per-image inference cost and latency scale unfavourably for production document IE relative to a local pipeline; sending document images to a third-party API introduces privacy and data-control concerns that a deployer would have to address; and the resulting extraction behaviour cannot be audited by any party other than the model provider. None of these are objections to vision-language models as a research direction; they are deployment considerations specific to the use pattern in which a frontier VLM serves as the extractor, and they are part of why the thesis's pipeline operates locally on open-weight components and deterministic logic.

In the terms of \S\ref{sec:thesis_position_supervised_regime}, OCR-free document models and frontier-VLM-based extraction are alternatives that remove the staging that makes modular extraction possible. The thesis instead retains the staging, accepting the costs \S\ref{sec:ocr_free_shift} named --- OCR error propagation, staging overhead, and information loss at the OCR boundary --- as the price of the diagnostic transparency and audit trail the modular pipeline provides.

% =============================================================================
\section{Graph-based key information extraction}
\label{sec:graph_kie}

\subsection{Graph-based KIE as a research line}
\label{sec:graph_kie_research_line}

\S\ref{sec:bg-graphs-intro} introduced graphs as a representation for document reasoning; this section situates that representation within the broader graph-based KIE literature. The framing of document IE as a \textit{relational} problem follows from the observation that serialised text representations alone do not capture the structural relationships that determine meaning in visually rich documents. A document encodes key--value pairs in which the key and the value may be on the same row or in adjacent regions; header--data attachments in which a column or section header governs the data tokens beneath it; and row--column groupings in which membership in a logical row is determined by horizontal alignment rather than reading-order proximity. Tokens that are visually adjacent or structurally related may be far apart in any natural reading order, and conversely tokens close in reading order may belong to unrelated parts of the document. The natural data structure for representing relational reasoning of this kind is the graph: nodes correspond to tokens or token spans after OCR has been applied, and edges encode pairwise relationships between them. Graph-based KIE is the research line that has organised itself around this representation.

The systems that constitute this research line share three architectural commitments. First, they take OCR tokens or token spans as the nodes of a document graph constructed at inference time. Second, they encode pairwise relationships through edges that may be \textit{typed}, weighted, or otherwise conditioned by the relation they represent --- spatial proximity, alignment along a row or column, predicted dependency, key--value linkage --- rather than treating all token-pair relationships as a single homogeneous adjacency. Third, they consume the resulting graph structure through graph-aware reasoning machinery: graph convolution, message passing across edges, or attention mechanisms that condition on edge information as well as on node features. Within these shared commitments, the systems differ on three axes that \S\ref{sec:graph_kie_systems} will use to compare them: how the edges are produced --- by a learned module trained on labelled examples or by deterministic rules over geometry and semantic role; how the graph is used downstream --- as input to a graph neural network whose output is a tagged sequence, or as a structure consumed by a matching procedure; and what supervision signal is required for training, including whether labelled target-schema examples are needed at all. This final axis is the one most important for this thesis: some systems learn edges or graph-aware representations from labelled documents, whereas this thesis constructs its graph deterministically from geometry, semantic labels, and domain constraints.

\subsection{Representative graph-based KIE systems: PICK, SPADE, and FormNet}
\label{sec:graph_kie_systems}

PICK~\cite{yu2020pick} is a supervised graph-learning system for key-information extraction from documents. The model takes OCR-recognised text segments together with their textual content, visual segment features, and 2D coordinates as input. Over these inputs, it learns a pairwise edge structure --- an adjacency representation in which the weight associated with each token-pair edge conditions on textual, visual, and layout features jointly --- so that the graph reflects relationships the model has been trained to find informative on labelled documents. The resulting weighted graph is then consumed by a graph convolutional network whose layers propagate information across edges and produce node-level contextual representations for each text segment. A downstream tagging head emits the key-information predictions, and the entire pipeline --- feature encoders, edge-weight learning, graph convolution, and tagging head --- is trained jointly end-to-end on labelled examples in the target schema. PICK's positioning in the field is as a system that brings graph learning together with graph convolution under a shared supervised objective, reducing reliance on manually defined edge rules by learning continuous edge weights from labelled data.

SPADE~\cite{hwang2021spade} formulates the same overall problem differently. It takes OCR tokens together with their 2D coordinates as input and treats document IE as \textit{spatial dependency parsing} over the token stream rather than as sequence tagging. The model predicts head--tail dependency edges between tokens together with relation labels --- including representation, grouping, and sub-grouping --- that name the kind of structural relationship each predicted edge stands for. The predicted edges are then decoded into a structured document graph through a procedure that resolves competing edge predictions, including the tail-collision-avoidance step already referenced in \S\ref{sec:bg-graphs}; the structured target output is recovered from this parsed dependency graph rather than from token-level tag predictions. SPADE is supervised end-to-end on labelled data in the target schema. It contrasts with PICK in two ways: the graph is produced by an edge-prediction module rather than as an adjacency learned to feed graph convolution, and it is consumed by a parsing-and-decoding procedure rather than by graph-aware feature propagation.

FormNet~\cite{lee2022formnet} responds to a different motivating problem within the same research line: the observation that OCR tokens cannot always be reliably serialised into the linear reading order that sequence-tagging models assume, particularly on dense forms where rows, columns, and side panels disrupt the dominant reading direction. FormNet's architectural response is structural augmentation of an otherwise transformer-based document model. \textit{Rich Attention} introduces an attention mechanism that conditions on the spatial relationships between tokens in addition to their content features, so that attention is computed in a layout-aware rather than purely content-aware manner. \textit{Super-Tokens} are graph-convolution-derived contextual tokens constructed from each token's neighbours before the transformer's self-attention layers operate, providing each position with a structurally-informed local context that compensates for the unreliable serial order. FormNet is trained and evaluated on form-understanding benchmarks such as FUNSD and CORD. Together, these systems show three ways graph structure enters supervised KIE: as a learned weighted graph consumed by graph convolution, as a predicted dependency graph decoded into structured output, and as graph-aware structural encoding inside a transformer-based document model.

\subsection{Where this thesis sits: deterministic graph association without supervised graph learning}
\label{sec:graph_kie_position}

The thesis's pipeline can be compared with the three \S\ref{sec:graph_kie_systems} systems along the same axes \S\ref{sec:graph_kie_research_line} introduced. On edge production, PICK, SPADE, and FormNet all produce edges through learned modules trained end-to-end on labelled documents --- PICK as an adjacency representation whose pairwise weights are fitted from data, SPADE as a head--tail dependency edge predicted token-by-token, FormNet as part of the structural augmentation that feeds its transformer encoder. The thesis instead constructs typed edges deterministically from geometric primitives and semantic role assignments --- SAME\_ROW, SAME\_COL, and CONTEXT\_SCOPE, with continuous weights derived from spatial distance in the local coordinate frame of each structural unit, as \S\ref{sec:bg-graphs} describes. On downstream graph use, the three systems consume their respective graphs through learned operators: PICK through graph convolution that propagates information across edges, SPADE through a parsing-and-decoding procedure that recovers the structured target from predicted dependencies, FormNet through Rich-Attention layers that condition transformer attention on graph-derived spatial encoding. The thesis consumes its graph through a hybrid matcher built from operators that have no learned parameters: greedy bipartite matching for typical cases, structural propagation for the one-to-many associations bipartite matching cannot cover, and a rank-under-context fallback for cases where geometric matching has failed. On supervision, all three systems are trained end-to-end on labelled examples in their respective target schemas, while the thesis uses no labelled target-schema data --- the entire graph construction, edge weighting, and matching machinery is determined by domain rules and geometric primitives.

What the thesis takes from this research line is its representational substrate: the typed-edge document graph, the principle that document IE benefits from graph reasoning rather than from sequence tagging alone, and specific architectural ideas including the soft-adjacency framing of PICK and the tail-collision avoidance of SPADE that \S\ref{sec:bg-graphs-detlearn} named at the primitive level. What the thesis leaves behind is the supervised learning machinery --- both the labelled-data dependency that learning continuous edge weights or dependency predictions requires, and the graph-to-output prediction heads that translate graph structure into structured output through trained parameters. In the terms of \S\ref{sec:thesis_position_supervised_regime}, graph-based KIE is the research line whose representation this thesis \textit{retains and re-implements without supervised training}: it keeps the graph-based view of document IE, but replaces supervised graph learning with deterministic construction and matching.

% =============================================================================
\section{Zero-shot information extraction and semantic matching}
\label{sec:zero_shot_ie}

\subsection{Zero-shot IE as a research line}
\label{sec:zero_shot_research_line}

Zero-shot information extraction, in the sense relevant to this chapter, is the family of methods that operate without labelled target-schema training data. The supervised regime of \S\ref{sec:supervised_paradigm} presupposes labelled examples in the target schema and adapts a pretrained model to them; the supervised graph-KIE systems of \S\ref{sec:graph_kie} depend on labelled examples for edge production and downstream graph use; the alternative paradigms of \S\ref{sec:tsr} and \S\ref{sec:vlm_ocr_free} either change the extraction target or remove the staging that makes modular extraction possible, but in their dominant form also depend on supervised training in some way. The zero-shot family is defined in opposition to that dependence. Its methods produce structured output without fitting parameters to labelled examples in the target schema, and their generalisation comes from prior knowledge brought into the inference setting from elsewhere. \S\ref{sec:bg-class-zeroshot} already named three of these prior-knowledge sources; this section restates the framing and extends it with a fourth.

The first face is \textit{rule-based zero-shot extraction}: no parameters are fitted from labelled data, and the prior knowledge is what the developer encodes in lexicons, regular expressions, and decision rules --- the route the thesis's default classifier takes. The second is \textit{schema-flexible learned NER}: a model trained on cross-domain NER data is applied to a new schema by supplying the entity types as natural-language prompts at inference time, with prior knowledge transferred from the training distribution to the new schema rather than fitted to it. The third is \textit{embedding-prototype matching}: a pre-trained multilingual sentence encoder produces vector representations of candidate strings and developer-written category prototypes, and classification is performed by cosine similarity, with prior knowledge drawn from the encoder's pretraining geometry. The fourth, which extends \S\ref{sec:bg-class-zeroshot}'s framing, is \textit{LLM-prompted text-to-structure extraction}: a generative language model trained on broad text-and-instruction data is prompted with a schema specification and asked to emit structured records directly from input text, with prior knowledge drawn from generative pretraining and instruction-tuning combined. The four faces correspond to four different sources of prior knowledge --- encoded rules, cross-task transfer, pretrained embedding geometry, and generative-LM instruction-following --- and the rest of this section uses that distinction to organise the comparison. The important point for this chapter is that all four operate without labelled examples in the target schema, but they make different bets about what kind of prior knowledge is most useful, and the choice between them is therefore a methodological commitment rather than a uniform technique.

\subsection{Representative zero-shot IE families}
\label{sec:zero_shot_families}

Because rule-based zero-shot extraction is the thesis's default mechanism and was described in \S\ref{sec:bg-class-rules} and \S\ref{sec:bg-class-zeroshot}, this subsection focuses on the three non-default zero-shot families against which that default is positioned.

The first non-default family is \textit{schema-flexible learned NER}, whose canonical representative is \textbf{GLiNER}~\cite{zaratiana2023gliner}. The model is a compact bidirectional transformer encoder trained on a broad corpus of supervised NER data, with the distinguishing architectural commitment that the target entity types are supplied at inference time as natural-language strings rather than as a fixed set of output classes baked into a classifier head. The encoder consumes the input text together with the entity-type prompts, and a span-classification head scores candidate spans against each prompted type; a span is committed to a type when its score exceeds a threshold and the type is the argmax over the candidates the model has been asked to consider. Because the entity types are passed in at inference, the schema can be changed by changing the prompts: a new entity type --- \textit{supplement nutrient name}, \textit{unit of measurement}, \textit{reference-base context} --- can be supplied without retraining. GLiNER and later schema-flexible NER variants share the same basic design commitment: the entity schema is supplied at inference time rather than fixed by a task-specific classifier head. The family's prior-knowledge source is \textit{cross-task transfer}: the entity-recognition task has been learned across many schemas during training, and the family's generalisation to a new schema rests on the assumption that the task transfers without requiring the new schema to have been seen.

The second non-default family is \textit{embedding-prototype matching}. Its representative system in this thesis is \textbf{BGE-M3}~\cite{chen2024bgem3}, a multilingual sentence encoder that produces dense vector representations across more than a hundred languages and combines dense, sparse, and multi-vector retrieval functionality at the encoder level. Within the family, classification is performed by encoding both the candidate input and a developer-written prototype pool through the encoder, computing cosine similarity between the candidate and each prototype, and assigning the candidate to the category whose closest prototype clears a similarity threshold and a margin against competing categories. The same matching pattern is older than BGE-M3 itself: \textbf{Sentence-BERT}~\cite{reimers2019sentencebert} established the use of Siamese sentence encoders for efficient semantic similarity, and the prototype-matching mechanism builds directly on that recipe. What distinguishes BGE-M3 within the family is its scale and multi-granularity rather than the matching mechanism itself. The family's prior-knowledge source is the \textit{pretrained embedding geometry} of the encoder: the assumption is that the encoder's pretraining distribution has produced a vector space in which semantically similar strings sit close together regardless of language, surface form, or schema, against which developer-written prototype pools serve as anchors for application-level categories.

The third non-default family is \textit{LLM-prompted text-to-structure extraction}. The family takes an instruction-tuned generative language model and prompts it with a schema specification together with the input text, asking the model to emit structured output --- typically JSON-formatted records conforming to the prompted schema --- directly. The model is not fine-tuned on the target schema; the schema is communicated at inference time through the prompt, and the model's behaviour relies on its instruction-following capability and on its general capacity to produce structured output in response to a structured request. Representative systems include open-weight families such as the \textbf{Qwen-2.5} family~\cite{yang2024qwen25} and the \textbf{Llama} family~\cite{touvron2023llama2}, together with proprietary systems such as the GPT-4 family and the Claude family, all of which can be prompted to extract structured records from OCR text given a sufficiently specified schema in the prompt. The family's prior-knowledge source is the combination of generative-language-model pretraining on broad text corpora and instruction-following capability acquired through supervised fine-tuning and reinforcement learning from human feedback in the general-LLM training pipeline. This discussion concerns LLMs used as extractors, in which the model itself produces the target structured records from input text. The use of LLMs as evaluators of records produced by another system --- the role assigned to the LLM judge in \S\ref{sec:bg-evaluation-twotier} --- belongs to evaluation methodology rather than to the related extraction paradigms surveyed here.

\subsection{Where this thesis sits}
\label{sec:zero_shot_position}

The thesis's engagement with the three non-default zero-shot families surveyed in \S\ref{sec:zero_shot_families} is differential: two were evaluated as candidate replacements for the rule-based default, the third was not adopted as an extraction mechanism. For schema-flexible learned NER, the thesis evaluated GLiNER as a candidate replacement for the rule-based classifier of \S\ref{sec:bg-class-rules}, treating it as a plausible alternative whose cross-task transfer might recover nutrient names, units, or context labels that lexicon-based rules miss; the rule-based default was retained. For embedding-prototype matching, the thesis evaluated BGE-M3-based prototype matching with the threshold and margin parameters introduced in \S\ref{sec:bg-embeddings} as an alternative classifier path, treating the encoder's pretrained embedding geometry as a candidate substitute for the rule-based decision rule; the rule-based default was retained. For LLM-prompted text-to-structure extraction, the thesis did not adopt the family as its extraction mechanism. Three reasons are sufficient at the related-work level: it weakens \textit{determinism}, because the output varies across runs, model versions, and prompt formulations; it weakens \textit{stage-level control}, because a single generation collapses OCR, classification, and association into one opaque step; and it weakens \textit{auditability}, because the four diagnostic CSVs of \S\ref{sec:bg-evaluation-diagnostics} have no analogue in a generative-LLM extraction pass. These concerns mirror the modularity and reproducibility concerns raised for frontier-VLM extractors in \S\ref{sec:why_explicit_ocr}, but here they apply to text-only LLM extraction after OCR rather than to image-based VLM extraction. LLMs are used in the evaluation pipeline as judges per \S\ref{sec:bg-evaluation-twotier}; they are not used as extractors. The empirical comparison among the rule-based default, GLiNER, and the embedding-prototype alternative is treated in the experimental chapter.

In the terms of \S\ref{sec:thesis_position_supervised_regime}, zero-shot information extraction is the research line that \textit{operates without labelled target-schema data} --- the third of the three structural responses \S\ref{sec:thesis_position_supervised_regime} enumerated. The thesis sits squarely within this research line: its rule-based default and its evaluated alternatives all satisfy the no-labelled-target-schema-data condition, with the variation across them being the source of prior knowledge each method draws on. What this subsection establishes about the thesis's choice is not that the rule-based default is intrinsically superior to the alternatives --- it is that, given the constraint set \S\ref{sec:bg-problem} and \S\ref{sec:thesis_position_supervised_regime} named (zero-shot operation, layout-flexibility within domain, modular and interpretable design), the rule-based default and the evaluated alternatives all fit the constraint set, while the LLM-prompted family does not. The thesis selects within the zero-shot research line the mechanism that best preserves modularity and deterministic control.

% =============================================================================
\section{Nutrition-label and food-data extraction}
\label{sec:nutrition_label_extraction}

\subsection{Prior work specifically on nutrition-label extraction}
\label{sec:nutrition_prior_work}

Compared with the benchmark-driven literature on forms, receipts, and tables, the literature on nutrition-label extraction is comparatively sparse and more application-oriented than benchmark-oriented. As \S\ref{sec:doc_ie_problem} noted, nutrition and supplement labels are not part of the canonical benchmark set around which document KIE has largely organised itself, and the literature on this domain is therefore smaller and less leaderboard-driven than the work surveyed in \S\ref{sec:supervised_paradigm} and \S\ref{sec:graph_kie}. The work that does exist tends to combine OCR with downstream rules or, more recently, with LLM-based post-processing, in pipelines targeted at specific extraction goals --- recovering nutrient names and values, normalising units, populating downstream databases --- rather than at general benchmark performance. A recent example is the work of Assiri et al. on extracting nutritional elements and values from bilingual food labels using large language models~\cite{assiri2025bilingual}, which reflects the broader trajectory in document IE toward OCR-and-LLM hybrids and applies that trajectory specifically to nutrition labels in a multilingual setting.

What existing nutrition-label extraction work typically targets is the recovery of visible nutrient names together with their associated quantities and, in many cases, unit normalisation; the extraction goal is the data displayed on the label, often with limited handling of multilingual surface forms or of the reference-base context under which a value is reported. The thesis's target schema is narrower in some respects and wider in others: narrower because it is restricted to supplement labels rather than general food labels, wider because it requires full $\langle$nutrient, quantity, unit, context$\rangle$ tuple association in which each numerical value must be paired with the correct unit \textit{and} the correct reference-base context --- per\_100g, per\_serving, per\_daily\_dose --- across the multilingual layouts that EU supplement labelling routinely produces. The association problem, in this stronger sense --- pairing each nutrient with the correct quantity, unit, and context across slash-rows, multi-context columns, and paragraph-style declarations --- appears not to have been a central evaluation target in the published nutrition-label work to date. Adjacent OCR-on-packaging work and food-product databases provide infrastructure that the supplement-label setting can draw on, but they are not direct precedent for the specific tuple-association task that defines this thesis; \S\ref{sec:nutrition_adjacent} takes them up next.

\subsection{Adjacent food-data and retail-packaging extraction}
\label{sec:nutrition_adjacent}

The most direct neighbour of nutrition-label extraction is OCR work on food packaging more generally. \textbf{HalalBench}~\cite{halalbench2026} is a recent open multilingual OCR benchmark for food-packaging ingredient lists, motivated explicitly by the lack of standard OCR benchmarks for food packaging and by the practical challenges the domain presents --- curved surfaces, dense multilingual text, and sub-8pt fonts of the kind supplement labels routinely use. Comparative OCR studies have evaluated production engines including Tesseract, EasyOCR, PaddleOCR, and TrOCR on real-world food-packaging images that include both ingredient lists and nutrition-facts panels~\cite{nagayi2025foodpackagingocr}, providing applied OCR-engine comparisons on this kind of imagery. The role of this work for the thesis is at the perceptual layer rather than at the extraction-target layer: these studies address the perceptual layer of food-data extraction, but not the downstream nutrient--quantity--unit--context association problem targeted here.

A second strand of adjacent work consists of food-product databases that organise nutrition data at scale without claiming to be complete extraction architectures. \textbf{Open Food Facts}~\cite{openfoodfacts} is a major crowd-contributed food-product and nutrition-data resource: a multilingual database of food products including ingredients, allergens, and nutrition facts, populated through a combination of crowd contribution, automatic extraction, and curation. The Open Food Facts project also maintains an active interest in automatic nutrition-facts extraction from photographs, which indicates that automatic extraction from nutrition-facts images is an active applied problem outside the academic literature. \textbf{FLIP}~\cite{ahmed2022flip} is an analogous database focused on Canadian packaged foods, providing label-data resources for downstream use. These databases support the domain motivation for nutrition-data extraction at a scale academic benchmarks do not yet cover, but they are data resources rather than extraction systems, and they do not constitute precedent for the specific zero-shot tuple-association task this thesis targets.

\subsection{The gap this thesis addresses}
\label{sec:nutrition_gap}

The first version of the gap is at the benchmark level. No widely adopted canonical benchmark comparable to FUNSD, SROIE, CORD, or PubTabNet exists for the specific supplement-label tuple-association task that this thesis targets. The consequence is that published nutrition-label work, including the recent OCR-and-LLM hybrids \S\ref{sec:nutrition_prior_work} surveyed and the food-packaging OCR studies \S\ref{sec:nutrition_adjacent} cited, evaluates against bespoke targets defined within each study rather than against a shared benchmark, which makes cross-system comparison difficult and the leaderboard-driven progress that \S\ref{sec:doc_ie_problem} described unavailable in the way it has been for forms, receipts, and tables.

The second and more substantive version of the gap is at the level of combined constraints. The gap is not that nutrition labels have been ignored, but that existing work rarely combines the four constraints this thesis places at the centre: \textit{zero-shot operation}, \textit{multilingual supplement-label vocabulary}, \textit{explicit nutrient--quantity--unit--context association under multiple reference bases}, and an \textit{interpretable modular pipeline whose intermediate stages can be inspected}. Each of the four constraints is addressed individually somewhere in the surveyed literature --- zero-shot operation in the research line surveyed in \S\ref{sec:zero_shot_ie}, multilingual OCR on packaging in the comparative work cited in \S\ref{sec:nutrition_adjacent}, the relational-association problem implicitly addressed by the graph-KIE family surveyed in \S\ref{sec:graph_kie}, and modular cascaded pipelines in the OCR-dependent paradigm surveyed in \S\ref{sec:ocr_free_shift} --- but their combined application to the supplement-label setting under EU labelling conventions is not directly supported by any prior system or benchmark the chapter has identified. The gap that this thesis addresses is therefore the combined application of these constraints to the supplement-label setting.

% =============================================================================
\section{Positioning of this thesis}
\label{sec:thesis_positioning}

The thesis's location across the three structural-response categories of \S\ref{sec:thesis_position_supervised_regime} can be stated as a configuration of choices. The thesis does not \textit{change the extraction target}: it does not adopt the table-structure-recognition path of \S\ref{sec:tsr}, which would reconstruct a grid before tuple extraction. Nor does it remove the staged OCR-to-reasoning architecture: it does not adopt the OCR-free or vision-language document-understanding path of \S\ref{sec:vlm_ocr_free}. The thesis \textit{retains the graph representation but re-implements it without supervised training}: it borrows architectural ideas from the graph-based KIE family of \S\ref{sec:graph_kie} --- typed-edge document graphs, soft adjacency, tail-collision avoidance --- and constructs them deterministically from geometric primitives and semantic role assignments rather than learning them from labelled data. The thesis \textit{operates without labelled target-schema data}: it sits within the zero-shot information-extraction research line of \S\ref{sec:zero_shot_ie}, with its rule-based default selected against GLiNER-style schema-flexible NER and BGE-M3-based embedding-prototype matching as evaluated alternatives, and with LLM-prompted text-to-structure extraction excluded as the extraction mechanism. This three-part restatement is the core summary of where the thesis sits in the surveyed landscape.

The same positioning can be stated as a response to the constraint set \S\ref{sec:bg-problem} introduced and \S\ref{sec:thesis_position_supervised_regime} restated: zero-shot operation, layout-flexibility within the supplement-label domain, and modular and interpretable design, with \S\ref{sec:nutrition_gap} adding the domain-specific commitment to nutrient--quantity--unit--context tuple association under multiple reference bases. Each of the surveyed alternatives, considered against this constraint set, leaves at least one of its constraints unresolved. TSR leaves the target-schema constraint unresolved, because it reconstructs a grid rather than the required tuples and would still require a downstream typing stage. OCR-free document models leave the zero-shot constraint unresolved, because their reliable schema-conformant operation depends on supervised adaptation to the target schema. Frontier-VLM extraction leaves the modularity and auditability constraints unresolved, because end-to-end generation hides the intermediate decisions a modular pipeline exposes. Supervised graph-KIE leaves the labelled-data constraint unresolved, because its edge production and graph-to-output prediction are fitted to labelled examples in the target schema. LLM-prompted text-to-structure extraction leaves determinism, stage-level control, and auditability unresolved for the same architectural reasons that frontier-VLM extraction does, applied at the text-only layer after OCR. The thesis's deterministic OCR-then-rules-then-graph configuration is not presented as universally strongest on every individual criterion; it is the configuration the chapter has identified as satisfying the constraint set as a whole.

This is the chapter's closing claim. The surveyed research lines have each contributed to one or more of the constraints in isolation, and the thesis borrows from several of them --- soft adjacency from PICK, tail-collision avoidance from SPADE, the broader graph-as-document-representation principle, multilingual sentence-encoder geometry for the embedding-classifier ablation, and the post-OCR repair tradition of cascaded document-IE pipelines --- but no single prior system surveyed in this chapter combines the four constraints in the configuration the supplement-label setting requires. The thesis's configuration is justified not by claiming universal superiority over the alternatives, each of which is stronger on its own optimisation axis, but by being the configuration this chapter has identified as fitting the four constraints in combination.