% !TEX root = thesis.tex

% =========================================================================
% Chapter 2 — Background
%
% Required packages (already loaded by thesis.tex):
%   inputenc (utf8), graphicx, amsmath/amssymb, booktabs, url,
%   hyperref, libertine.
%
% Bibliography keys required (in references.bib):
%   long2021scene, baek2019craft, liao2020db, zhou2017east,
%   shi2017crnn, graves2006ctc, sheng2019nrtr, du2022svtr,
%   devlin2019bert, zaratiana2023gliner, yu2020pick, hwang2021spade,
%   kuhn1955hungarian, mikolov2013word2vec, pennington2014glove,
%   reimers2019sentencebert, chen2024bgem3
%
% Search before submission for "[PLACEHOLDER" to find the LLM-evaluator
% example that needs swapping in §2.7.4.
% =========================================================================

\chapter{Background}\label{ch:background}

This chapter establishes the technical foundation that the rest of the thesis builds on. The pipeline of this thesis transforms images of supplement nutrition labels into structured records, and that transformation depends on a specific stack of concepts and primitives --- optical character recognition, document geometry, token classification, graph reasoning, semantic embeddings, and a quantitative evaluation regime --- each of which has its own history in the document-understanding and natural-language-processing literatures. The chapter introduces these concepts at the level of detail required to follow the methodology and experimental chapters that come next, in the order in which the pipeline encounters them. Positioning of the thesis against specific published systems and against the broader research lineage is the work of Chapter~\ref{ch:related-work}; Chapter~\ref{ch:background} confines itself to the conceptual machinery the pipeline depends on.

The chapter is organised as follows. \S\ref{sec:bg-problem} frames the extraction-and-association problem the pipeline solves and states the three commitments --- zero-shot, layout-flexible within the supplement-label domain, and modular and interpretable --- that the rest of the thesis is structured around. \S\ref{sec:bg-ocr} covers optical character recognition and the post-OCR repair stage that follows. \S\ref{sec:bg-layout} introduces the geometric primitives --- bounding boxes, columns, rows, headers, and the formalism the pipeline uses to handle column slope. \S\ref{sec:bg-classification} surveys the three classifier families the thesis evaluated as candidate implementations of the classification stage. \S\ref{sec:bg-graphs} develops the graph representation that the association stage operates on, including the typed edges, the tail-collision-avoidance step, the bipartite matching procedure, and the rank-under-context fallback. \S\ref{sec:bg-embeddings} deepens the treatment of embedding-based classification specifically, introducing the multilingual sentence encoder used in the embedding ablation. \S\ref{sec:bg-evaluation} defines the evaluation regime, the four-field tuple metric, and the two-tier matching protocol the experimental chapter relies on. \S\ref{sec:bg-philosophy} closes by articulating the architectural philosophy that runs through the chapter as a whole.

% =========================================================================
\section{Problem Framing and Pipeline Commitments}\label{sec:bg-problem}
% =========================================================================

The pipeline of this thesis takes an image of a supplement nutrition label as input and produces structured tuples as output. The label may be presented in any of the layouts encountered in the European supplement market: a flat table, a multi-column panel, a skewed or curved label, or a paragraph-style declaration in which the values are embedded in flowing prose. The pipeline must solve two problems jointly. The first is \emph{extraction}: recovering nutrient names, numeric values, units of measurement, and reference-base contexts from the image text. The second is \emph{association}: determining which nutrient each value belongs to, which unit accompanies which value, and which reference base each pairing is reported under. The output is a structured record of the form $\langle \text{nutrient},\,\text{quantity},\,\text{unit},\,\text{context} \rangle$, where the \emph{nutrient} names what is being measured, the \emph{quantity} and \emph{unit} give the numerical value and its unit of measurement, and the \emph{context} names the reference base on which the value is stated (per~100~g, per serving, per daily dose). A tuple counts as correct only when extraction has succeeded for all four fields and association has linked them to the same underlying fact on the label. The pipeline is built for this domain --- it is not a general document extraction system --- but within the domain it is designed to handle the full range of layouts that commercial supplement labelling presents.

Supplement labels are visually small, information-dense, and idiosyncratic, and the combination of these properties makes them a harder extraction target than the receipts and forms that dominate the document-AI literature. Five properties recur. \textbf{Dense numeric content}: a typical label encodes between eight and forty quantitative facts in a single rectangular region a few square centimetres in size, and adjacent rows differ only by a single nutrient name and a small number, so a one-row error in association destroys the tuple. \textbf{Multi-context columns}: EU labels frequently report nutrients twice in adjacent columns, once per~100~g and once per serving, and the two columns of identical numbers must be distinguished only by their column header, which sits several rows away from the data it governs. \textbf{Multilinguality}: EU directive 1169/2011 permits labels in multiple official languages on the same package, and slash-separated polyglot rows such as \emph{Eiwei{\ss} / Protein / Prot{\'e}ines} appear routinely on products sold across the European market. \textbf{OCR-hostile typography}: labels are printed in small fonts (often below 6~pt for non-mandatory information) on curved or wrinkled plastic surfaces, and feature dense table borders that confuse line-detection heuristics. \textbf{Sub-nutrient hierarchy}: rows such as \emph{davon ges{\"a}ttigte Fetts{\"a}uren} (of which: saturated fats) are conceptually a sub-component of the row above them, but they appear at the same visual indentation as their parent, and a pipeline that reasons only on visual layout will treat them as independent peer nutrients rather than as nested under their parent. A sixth and qualitatively different case is \textbf{non-tabular labels}, in which the values are not arranged in a grid at all but are embedded in flowing prose: ``Per~100~g: 1500~kJ energy, 12~g protein, 45~g carbohydrates of which 8~g sugars, \ldots''. A pipeline restricted to tabular reasoning fails entirely on these labels.

The first commitment is \textbf{zero-shot}. The pipeline does not use labelled examples of the target task to train, fit, or tune any component. Lexicons, rules, and decision logic are written from domain knowledge --- the EU 1169/2011 nutrient-name vocabulary, the standard units of measurement, the canonical reference-base context labels --- before any image is processed. Labelled data is used only for evaluation, to measure how well the pipeline performs against ground truth, and is never consulted during inference. The single exception is threshold selection for the embedding-based classifier, where a held-out subset of labelled data is used to choose decision thresholds via grid search; this follows standard practice for prototype-based zero-shot classifiers and does not constitute training in the conventional sense, since no model parameters are fit. The cost of this commitment is that the pipeline cannot learn from its own mistakes --- every failure mode must be diagnosed by hand and addressed by changing rules, lexicons, or geometric primitives, with no gradient signal to absorb the error. The benefit is that the pipeline generalises to any supplement label without re-annotation: a new product, a new layout, a new language combination is handled by the same code that handled the labelled evaluation set, with no retraining cycle. This generalisation is what makes the pipeline deployable at the scale of the European supplement market, where tens of thousands of products with new layouts enter circulation each year.

The second commitment is that the pipeline must \textbf{work across the full range of layouts} that supplement labels exhibit, without layout-specific code paths. Three families of mechanisms deliver this commitment together. \emph{Geometric mechanisms} --- adaptive row thresholding, column-derived direction inference, and column-role inference --- handle the typical case in which the layout is recoverable from local geometry. \emph{Structural mechanisms} --- rank-under-context row membership and translation-row collapsing --- recover cases where local geometry is unreliable, by reasoning about which token holds which structural position within its column rather than by measuring pixel distances. \emph{Surface-form fallback} --- a paragraph-level extractor --- handles the case where no tabular structure is recoverable at all, taking over when row induction fails or when the structured path produces no output. The cost of this commitment is that no single mechanism can be tuned to optimum: each must work well enough on the inputs it sees so that the family below it is not invoked unnecessarily, and the architecture is more involved than a layout-specific pipeline would be. The benefit is that the pipeline does not fail catastrophically on unfamiliar layouts: a curved label, a multi-column panel, a paragraph-style declaration, or a label with unusual row spacing is handled by the same code that handles the typical case, because the typical case is just one point in a continuum the pipeline already covers.

The third commitment is that the pipeline must be \textbf{modular and interpretable}. The pipeline is organised as a sequence of independent stages: optical character recognition, post-OCR correction, semantic classification and token enrichment, graph construction, tuple association, and evaluation. Each stage has a typed input--output contract: it receives a fixed schema from the stage above it and emits a fixed schema to the stage below, and no stage depends on the internal implementation of another. This permits, for example, swapping the OCR engine without touching the classifier, or replacing the rule-based classifier with an embedding-based one while keeping the graph unchanged. Each stage also emits a diagnostic record that captures its internal decisions: how many tokens it received, how many it dropped, which correction rules fired, which classifier fallback was used, what edges the graph constructor pruned, why a given association was made. The diagnostic records concatenate into a per-image audit trail in which every output tuple can be traced backwards to the OCR tokens that produced it. The cost of this commitment is engineering overhead: each stage must define and uphold its schema, the diagnostic logging adds to the codebase, and a single global optimisation across stages is not available because the stages cannot reach into one another. The benefit is that failure modes are localisable: a regression on a single image can be attributed to a specific stage, a specific token, and a specific decision, which is what makes single-variable ablations a viable methodology and what allows the experimental chapter to make stage-level causal claims of the form ``swapping the OCR engine raised nutrient F1 by $N$ percentage points'' with no other change in the pipeline.

% =========================================================================
\section{Optical Character Recognition and Post-OCR Repair}\label{sec:bg-ocr}
% =========================================================================

\subsection{OCR as the Perceptual Entry Point}\label{sec:bg-ocr-entry}

Optical Character Recognition is the perceptual entry point of the pipeline. It is the only stage that operates on pixel data: every later stage consumes its output and never sees the original image. This places OCR in a distinctive position relative to the rest of the pipeline. Recognition errors at this stage are unrecoverable in a strict sense --- no downstream component, however carefully engineered, can reconstruct a token that OCR failed to detect. A misread numeric value propagates through the graph unchanged, and a missing nutrient name simply does not appear in the output. The perceptual layer therefore sets the ceiling on what every subsequent stage can achieve, which is why the choice of OCR engine and the post-OCR repair stage that immediately follows are treated as primary determinants of pipeline performance rather than as routine preprocessing.

\subsection{The Two-Stage Paradigm}\label{sec:bg-ocr-twostage}

Modern OCR pipelines often decompose text extraction into two cascaded sub-problems: text detection and text recognition \cite{long2021scene}. The detection stage takes a document image and produces a set of polygons --- usually quadrilaterals or axis-aligned rectangles --- that localise text-bearing regions. The recognition stage then takes each cropped region and transcribes it into a string. The factorisation is attractive for two reasons. It permits independent optimisation of the spatial and the lexical components, and the per-region crop removes most of the layout variability before recognition begins, since the recogniser sees often one word or line crop at a time rather than a whole page.

Three influential families of detection architectures are \textbf{CRAFT} \cite{baek2019craft}, which predicts a per-pixel region score together with an affinity score between adjacent characters and produces character-level boxes that can be merged into word-level polygons; \textbf{DBNet} \cite{liao2020db}, which integrates the binarisation step directly into the network so that the threshold itself is a learned function of position, giving sharper segmentation maps on dense text; and \textbf{EAST} \cite{zhou2017east}, which regresses quadrilaterals from a single fully convolutional network in one shot and supports rotated boxes, trading some boundary precision for substantially faster inference. Other detectors used in production OCR systems take different approaches --- progressive scale expansion (PSENet), pixel aggregation (PAN), transformer-based regression --- but the three above are the most influential references in the line that leads to the OCR engines this thesis evaluates. The choice between detector families primarily trades off polygon quality, robustness to curvature, and speed; supplement labels, with their dense rows and small fonts, benefit most from the higher-precision boundaries of CRAFT and DBNet.

Recognition transduces a cropped region image into a character string. The dominant architecture for almost a decade has been the Convolutional Recurrent Neural Network (\textbf{CRNN}) \cite{shi2017crnn}: a CNN encoder feeds a bidirectional LSTM, which feeds a Connectionist Temporal Classification (CTC) head \cite{graves2006ctc} that aligns variable-length predictions with the input without requiring per-character supervision. The CTC loss marginalises over possible alignments by introducing a special blank symbol and a many-to-one mapping that collapses repeated symbols and removes blanks; this marginalisation makes CTC robust to local uncertainty in character alignment, but it does not explicitly model two-dimensional layout or long-range semantic structure, and so CRNN-class recognisers can become fragile when the upstream detector provides crops that contain multi-line, multilingual, or table-like content. More recent recognisers replace the BiLSTM with attention decoders or with transformer-based modules --- examples include the Neural-machine-translation-style Reading Transformer (NRTR) \cite{sheng2019nrtr}, which uses a self-attention encoder--decoder without recurrence, and the Single Visual model for scene Text Recognition (SVTR) \cite{du2022svtr}, which abandons the recurrent module entirely in favour of patch-wise visual mixing. Hybrid recognisers run an attention branch and a CTC branch in parallel and either ensemble or distil between them. These attention-based recognisers can handle longer-range dependencies more flexibly than pure CTC, although multi-line document crops still require appropriate detection, rectification, or layout-aware modelling at the surrounding stages of the pipeline.

\subsection{OCR Error Taxonomy}\label{sec:bg-ocr-errors}

The output of an OCR engine on supplement labels is not error-free text but text with a characteristic distribution of errors, and the post-OCR repair stage that follows is designed against this distribution. Five categories recur across engines and across labels.

\textbf{Character confusion} is the substitution of one character for another with similar shape. The canonical examples are \texttt{0} for \texttt{O}, \texttt{l} for \texttt{1}, \texttt{rn} for \texttt{m}, and $\mu$ for \texttt{u}, the last of which is particularly damaging on the unit field because $\mu$g (micrograms) and mg (milligrams) differ by three orders of magnitude. Character confusion is most damaging on numeric tokens, where a single substitution flips the value, and on unit tokens, where the unit dictates the magnitude of the entire fact.

\textbf{Token fusion} is the failure to insert a whitespace boundary between adjacent units of meaning. A label printed as \texttt{400 mg} may be returned as the single token \texttt{400mg} when the detection stage produces a polygon that spans both the value and the unit. The downstream consequence is severe: the classifier may see one ambiguous token, the unit is never extracted as a separate node, and the tuple loses its unit field entirely.

\textbf{Row merging} is the dual failure: the detection stage groups two visually adjacent rows into a single text line, and the recogniser transcribes them as one string. The result is a token like \texttt{Eiwei{\ss} 12 g Fett 8 g} in which two complete nutrient--quantity--unit triples appear in a single OCR cell. A purely lexical rule can split such a token into substrings, but it cannot reliably recover which substring belongs to which row of the original label without spatial information from the bounding boxes. The row-merging case is one of the strongest arguments for a graph-based association stage that operates on geometry rather than on the raw token list.

\textbf{Non-nutritional text noise} arises when the detection stage produces line crops that span an ingredient list, a marketing claim, a regulatory disclaimer, or other prose unrelated to the nutritional declaration. The crop may also contain a mixture of languages, since EU labels are routinely multilingual. The recogniser receives input that is poorly aligned with the target extraction task and produces a low-confidence transcription with garbled segments and dropped tokens. The downstream effect is that some images return a token list dominated by NOISE-classified tokens with no recoverable nutritional content; these are the OCR-failure cases the experimental chapter discusses.

\textbf{Artefact injection} is the appearance of stray punctuation or formatting characters at the boundaries of otherwise correct tokens. Common artefacts include leading single quotes or square brackets, trailing pipes, and vertical bars used as table-cell separators that the recogniser misreads as \texttt{|} or \texttt{\}} characters; these prevent lexicon matching even when the underlying nutrient name is recognised correctly. The corrector strips them as a deterministic preprocessing step before the classifier sees the token stream.

\subsection{Post-OCR Correction and Text Normalisation}\label{sec:bg-ocr-correction}

Two deterministic text-transformation layers sit between the OCR output and the rest of the pipeline. The first is \textbf{post-OCR correction}, which attempts to repair the errors of \S\ref{sec:bg-ocr-errors}. The second is \textbf{text normalisation}, which maps valid surface forms to canonical representations. The two layers are conceptually distinct: post-OCR correction targets likely OCR-induced errors, while normalisation maps valid surface variation to canonical forms. Both operate by deterministic rules rather than learned parameters, but their inputs and outputs and their reasons for existing are different.

Post-OCR correction in this thesis is implemented as an ordered sequence of rules applied to the raw OCR token list. The rules target the failure categories from \S\ref{sec:bg-ocr-errors} directly: token-fusion rules split a single OCR token into multiple tokens when the token contains both a numeric value and an alphabetic unit (\texttt{400mg} is split into \texttt{400} and \texttt{mg}); regulatory-string rules strip non-nutritional substrings such as Nutrient Reference Value percentages that recur in standard EU label templates and would otherwise enter the classifier as spurious nutrient or context candidates; artefact-stripping rules remove leading and trailing punctuation that prevents lexicon matching. A small number of rules also merge tokens when an OCR-introduced whitespace has separated a name that should be interpreted as one semantic unit; in such cases the merged token is given a combined bounding box that is the union of its constituents, so that downstream stages reasoning about geometry continue to receive a single coherent spatial object. Each rule is a small deterministic transformation derived from inspection of representative failure cases on a development subset, and the rule set is extended by adding new rules rather than by retraining a model. The corrector emits a per-image log of which rules fired on which tokens, which feeds the diagnostic audit trail of \S\ref{sec:bg-evaluation-diagnostics}.

Text normalisation operates at two distinct points in the pipeline, with different roles. The first is at \textbf{inference time}, inside the classifier, where surface variants of the canonical context labels are mapped onto a closed set of canonical values: strings such as \texttt{PRO PORTION}, \texttt{Je 100g}, \texttt{pro Portion}, and \texttt{per 100 g} are collapsed onto canonical labels including \texttt{per\_100g}, \texttt{per\_100ml}, \texttt{per\_serving}, and \texttt{per\_daily\_dose}. This mapping is necessary for the pipeline to produce stable context-field values across labels written in different languages and styles, and it is consulted by the classifier on every token classified as a context candidate. The second is at \textbf{evaluation time}, inside the matching predicates of \S\ref{sec:bg-evaluation-tuplef1}, where additional normalisations are applied to the predicted and ground-truth strings before comparison: lowercase folding, German umlaut transliteration, diacritic stripping, decimal-comma standardisation. The two normalisation steps are deliberately separate. Inference-time normalisation produces the canonical output the pipeline commits to. Evaluation-time normalisation is a comparison aid that allows two strings already correct in their respective canonical forms to be recognised as equal despite incidental representational differences.

Keeping post-OCR correction and text normalisation as separate layers serves two purposes. The boundary between them is the boundary between repairing likely OCR errors and standardising valid surface variation, and a pipeline that conflates the two tends to over-correct: a single normalisation rule applied indiscriminately to OCR output may silently rewrite genuinely incorrect tokens into plausible-looking but wrong canonical forms. Keeping the layers distinct also keeps each one diagnosable. The corrector log records which OCR errors were repaired and which slipped through; the inference-time normaliser records which surface variants were canonicalised and which were left unrecognised; the evaluation-time normaliser is part of the metric definition, not part of the extraction pipeline itself. A failure in any one of these can be traced to its layer without disturbing the others.

% =========================================================================
\section{Document Layout and Geometric Structure}\label{sec:bg-layout}
% =========================================================================

\subsection{Bounding Boxes and Reading Order}\label{sec:bg-layout-bboxes}

After OCR and post-OCR repair, the pipeline represents each token as text plus geometry. The geometry locates the token on the page. OCR engines emit text-bearing regions as polygons --- typically quadrilaterals --- but the pipeline of this thesis approximates each polygon as an axis-aligned rectangle, represented by the image-pixel coordinates of its top-left corner $(x_1, y_1)$ and its bottom-right corner $(x_2, y_2)$, together with a centroid $(c_x, c_y)$ at the geometric centre of the rectangle. The axis-aligned approximation is sufficient for supplement labels, on which text is rarely strongly rotated and the precision lost by ignoring small in-plane rotations does not affect the downstream graph reasoning. The bounding box, in this representation, is the only spatial information any downstream stage receives. Once OCR has run, the original image is not consulted by the geometric, classification, graph, or association stages.

Among the things a bounding box does not encode is the \textbf{reading order} of the document. Reading order is the sequence in which a human would consume the tokens --- left-to-right within each row, top-to-bottom across rows, with column breaks honoured where columns exist. OCR engines do not always return tokens in reliable reading order; they may return them in detector order, in internal processing order, or in an approximate layout order that is not guaranteed to match the order a human reader would follow. On flat single-column documents the engine's output and the human reading order coincide closely, but on multi-column labels and on labels whose rows have any vertical curvature, they diverge: the engine's stream may interleave tokens from the left column with tokens from the right column, place column headers below the data they govern, or break a single visual row across two non-adjacent positions in the stream. Recovering reading order from the bounding-box stream is therefore a structural-inference problem rather than a transcription one, and the geometric mechanisms introduced in the rest of this section are the primitives the pipeline uses to solve it.

\subsection{Columns, Rows, and Headers}\label{sec:bg-layout-cols}

A bounding box locates a single token, but the meaning of a supplement label is carried by groups of tokens arranged into rows and columns. Neither rows nor columns are explicitly marked in the OCR output; both must be recovered from the bounding-box stream by reasoning about the spatial relationships between tokens. The pipeline therefore treats columns and rows as \textbf{inferred structures} --- derived from the bounding boxes rather than read off them --- and the rest of \S\ref{sec:bg-layout} is concerned with the primitives this inference uses.

A \textbf{column} is a vertical band of the label in which a single semantic role is repeated down the page: a column of nutrient names, a column of per-100g quantities, a column of per-serving quantities, a column of percentage values. In the ideal case of a flat label, all tokens belonging to the same column share a near-constant $c_x$, and column membership is decided by simple ranges over $c_x$. In practice, the pipeline infers column membership approximately, since labels exhibit local skew and curvature that cause $c_x$ to drift down a column. A \textbf{row} is the horizontal counterpart: a band in which one nutrient's full record --- name, quantity, unit, context, and any auxiliary fields --- is reported across the columns. Rows are locally horizontal rather than globally aligned: on flat labels they coincide with constant $c_y$ ranges, but on curved or skewed labels their orientation drifts across the page. The inference of column and row membership under these distortions is the central topic of the geometric and graph stages of the pipeline.

A \textbf{header} is a special class of row whose tokens label the columns rather than carry data values. Most supplement labels have one header row, near the top of the table, that names each column with a context phrase such as \texttt{per 100g}, \texttt{per serving}, or \texttt{\% NRV}. Some labels have two header rows --- a primary header naming the columns and a secondary header giving units or percentage references. A header row differs from a data row in two ways: its tokens are typically classified as CONTEXT, UNIT, or as percentage-related tokens rather than as NUTRIENT or QUANTITY, and the column-role inference of the pipeline (introduced in \S\ref{sec:bg-graphs}) treats header tokens as the column's canonical descriptor rather than as nutrient records. Identifying which row of a label is the header is itself an inference task, and the absence of a unique header signal --- no font weight, no separator line that survives OCR reliably --- is what forces the pipeline to identify headers from the semantic labels of their constituent tokens rather than from any visual cue.

\subsection{Geometric Distortions}\label{sec:bg-layout-distortions}

The flat-label assumption that simplifies column and row inference rarely holds in practice. Supplement labels are printed on physical packaging --- bottles, jars, sachets, blister packs --- and a label photographed in commercial conditions exhibits one or more of the following geometric distortions.

\textbf{Skew} is a small global rotation of the label about its centre. A label that should sit at zero degrees instead sits at, say, three degrees clockwise, so that what was a horizontal row of tokens becomes a slightly inclined line. Under the pipeline's axis-aligned bounding-box representation, the boxes themselves remain axis-aligned, but the token centroids of a logical row no longer share a common $c_y$ value --- they lie along an inclined line rather than a horizontal one. The principal effect on the pipeline is that the constant-$c_y$ assumption for row membership breaks, and row inference must compensate.

\textbf{Rotation} is the more general case in which the label is rotated by an arbitrary angle, possibly large. A photograph of a bottle taken from below or held sideways can produce a label rotated by ninety degrees or more. Rotation is, mathematically, a generalisation of skew, but is treated separately because the magnitude matters: small rotations can be absorbed by the row-inference mechanism, while large rotations require a preprocessing step that re-orients the image before OCR is run. The latter case is outside the scope of the pipeline of this thesis, which assumes that input images are roughly upright.

\textbf{Curvature} is a local rather than global distortion. A label printed on a cylindrical bottle, photographed from the front, exhibits a horizontal compression at the edges of the label that no global rotation can describe. Two consequences follow. First, rows in different vertical regions of the label, or even different segments of the same row, may have different local orientations: a row near the top of a curved label may be near-horizontal while a row near the bottom is inclined by several degrees, and a single long row that spans a curved region may itself bend across its width. Second, the relationship between $c_x$ and column membership is no longer linear: a column may drift sideways across the page as it descends, so that the constant-$c_x$ assumption for column membership breaks in addition to the constant-$c_y$ assumption for rows.

A common preprocessing strategy in the document-IE literature is to estimate a single global skew angle and to rotate the image to remove it before OCR is run. This works well when the distortion is global. When the distortion is local --- as on curved or perspective-warped labels --- global deskewing cannot recover row alignment, because there is no single rotation angle that flattens the page. The pipeline of this thesis therefore does not rely on a deskewing preprocessor and instead handles geometric distortion through the row, column, and direction-inference primitives developed in \S\ref{sec:bg-layout-slope} and \S\ref{sec:bg-graphs}. The cost of this choice is methodological complexity: the inference primitives must be robust to the distortions that a deskew step would otherwise have eliminated. The benefit is that the pipeline does not break on labels for which no global deskew exists --- the curved and locally-warped cases that the dataset contains.

\subsection{Context-Header Attachment Under Column Slope}\label{sec:bg-layout-slope}

A particular failure mode arises on multi-context labels when the columns themselves are not perfectly vertical. EU labels routinely report nutrients twice in adjacent columns --- once per~100~g and once per serving, or once per~100~g and once per daily dose --- and each data row contains a quantity in each column, both governed by their respective column header. Each data quantity must be attached to the correct header so that the predicted tuple records the correct context. On flat labels this attachment is straightforward: the data token at horizontal position $c_x$ is attached to whichever header sits directly above it, where ``directly above'' means ``smallest $|c_x^{\text{data}} - c_x^{\text{header}}|$''. A simple horizontal-distance comparison is sufficient.

This breaks when the columns slope. Consider a multi-context label printed on a slightly skewed surface, or photographed at a small angle, so that the per-100g column drifts a few dozen pixels to the right as the table descends. At the top of the table, the per-100g header sits at $c_x = 200$ and the per-serving header at $c_x = 320$. At the bottom of the table, the per-100g column has drifted sideways and a per-100g data token now sits at $c_x = 280$. The horizontal distance from this token to its own header is $|280 - 200| = 80$ pixels, and the horizontal distance to the per-serving header is only $|280 - 320| = 40$ pixels --- half as far. The naive horizontal-distance heuristic attaches the token to the per-serving header instead of the per-100g header, the predicted tuple records the wrong context, and the 4F~F1 metric records the failure.

The fix is to attach data tokens to headers via the \textbf{direction the column itself runs in}, not via horizontal distance to the headers' positions at the top of the table. Each column's direction can be estimated from the line of best fit through the centroids of the tokens that have been preliminarily assigned to that column --- for example, by an initial coarse $c_x$-based grouping, or by restricting the fit to high-confidence column members whose membership is unambiguous from local geometry alone. Because columns on a supplement label are approximately vertical, the appropriate parameterisation is $x = a_k\,y + b_k$, in which $y$ (the vertical position) is the independent variable and $x$ (the horizontal position) is the predicted column position at that height. This form is numerically stable for nearly-vertical columns, where the alternative $y = m_k\,x + b_k$ form would have an unbounded slope. Given the fit, the perpendicular distance from a data token at position $p = (c_x^{\text{data}}, c_y^{\text{data}})$ to column $k$'s line is
\begin{equation*}
  d_k(p) \;=\; \frac{\left| c_x^{\text{data}} - a_k\,c_y^{\text{data}} - b_k \right|}{\sqrt{1 + a_k^{2}}},
\end{equation*}
obtained by rewriting the line in implicit form and applying the standard point-to-line distance formula. The data token is attached to the column $k$ --- and therefore to column $k$'s header --- that minimises $d_k(p)$, rather than to the column whose header has the smallest horizontal-distance value $|c_x^{\text{data}} - c_x^{\text{header}}|$.

This formulation generalises the flat-label and slope-affected cases under a single mechanism. On perfectly flat labels, every column's best-fit line is vertical (slope $a_k = 0$ in the $x = a_k\,y + b_k$ form) and the perpendicular-distance criterion reduces to horizontal-distance matching, so no special-case code is needed. On skewed labels, each column's best-fit line is inclined and the perpendicular-distance criterion accounts for the slope, so a data token whose horizontal position has drifted sideways is still attached to its proper column. On curved labels, the linear fit is an approximation rather than an exact description of the column path; whether the approximation is sufficient to recover the correct attachment is an empirical question that the experimental chapter examines image by image. The mechanism is one specific instance of the broader pattern that the geometric stages of the pipeline implement: instead of measuring static distances in image-pixel coordinates, the pipeline first estimates the \emph{local} coordinate frame of each structural unit (here, the column) and then measures distances in that frame.

% =========================================================================
\section{Token Classification and Named Entity Recognition}\label{sec:bg-classification}
% =========================================================================

\subsection{The Classifier-Stage Design Space}\label{sec:bg-class-design}

The classification stage of the pipeline assigns a semantic category to each OCR token or token span after the post-OCR repair stage has finished. The categories are a small fixed set: NUTRIENT, QUANTITY, UNIT, CONTEXT, NOISE, and UNKNOWN. NUTRIENT names what is being measured; QUANTITY and UNIT carry the value and its unit; CONTEXT records the reference base under which the value is reported; NOISE marks tokens that are part of the label but are not part of any nutritional fact; UNKNOWN marks tokens that resist all the classifier's rules. Every downstream stage of the pipeline --- graph construction, association, evaluation --- depends on the classifier having assigned correct labels to as many tokens or spans as possible.

Three families of approaches to this kind of token classification exist in the document-IE and natural-language-processing literature, and the thesis evaluated all three as candidate implementations of the classification stage. The first is \textbf{rule-based classification}, in which the classifier is a hand-engineered set of lexicons, regular expressions, and decision rules that the developer writes before any image is processed. The second is \textbf{zero-shot named-entity recognition} in the GLiNER style, in which a neural model trained on labelled NER data from other domains is asked at inference time to label spans on the supplement-label tokens, with the entity types specified as natural-language strings rather than as a fixed schema fitted on this data. The third is \textbf{embedding-based prototype matching}, in which a pre-trained multilingual sentence encoder produces vector representations of each candidate token, phrase, or normalised string, and compares them against vector representations of hand-written category prototypes by cosine similarity. The remainder of \S\ref{sec:bg-classification} introduces each family in turn, and \S\ref{sec:bg-class-zeroshot} contrasts them along the dimension that matters most for this thesis: the kind of prior knowledge each one relies on to achieve its zero-shot capability.

\subsection{Rule-Based Classification}\label{sec:bg-class-rules}

The simplest of the three classifier families is the \textbf{rule-based classifier}, in which classification decisions are made by deterministic rules written by the developer from domain knowledge. The classifier consists of three kinds of components: \textbf{lexicons} that enumerate the canonical surface forms of each category (a list of nutrient names in the languages the labels appear in, a list of unit symbols, a list of context surface variants), \textbf{regular expressions} that match patterns the lexicons cannot enumerate (numeric quantities, percentage values, NRV strings), and \textbf{decision rules} that combine the lexicon and regex matches with positional or contextual cues to commit to a final label. In this implementation, rules are evaluated in a fixed priority order, and the first rule that fires determines the token's label.

The strengths of the rule-based approach are interpretability and the strong sense of zero-shot. Every classification decision is traceable to a specific rule that the developer wrote and can inspect; when a token is labelled incorrectly, the responsible rule can be identified, modified, or supplemented without retraining anything. The approach also satisfies the strictest reading of zero-shot: no learned parameters exist anywhere in the classifier, no labelled training data is used to fit parameters, and the only domain knowledge brought to bear is whatever the developer has written into the lexicons and rules. The classifier is also computationally inexpensive, since each token is processed by a small fixed sequence of lexicon lookups and regex tests.

Rule-based classification is also a natural fit for a pipeline whose downstream stages depend on the \emph{stability} of category labels rather than on rich semantic generalisation. The graph constructor of \S\ref{sec:bg-graphs-edges} reasons over edge types defined by category labels, and the association stage of \S\ref{sec:bg-graphs-matching} propagates context through column-role inference. Both stages require that a token labelled NUTRIENT in one image be labelled NUTRIENT in another by the same logic, so that the graph operations are comparable across images. A rule-based classifier provides this stability by construction: identical inputs produce identical labels, with no run-to-run variability and no dependence on which images the classifier has been exposed to.

The limitations of the rule-based approach are also direct consequences of its construction. \textbf{Lexicon coverage} is the principal one: a token whose surface form is not in the lexicon is classified as UNKNOWN regardless of how clearly it represents a nutrient or unit to a human reader. New product categories, new vitamin shorthand, OCR-induced surface variants, and unusual translations all require manual lexicon extension. \textbf{Surface-form variation within categories} is a related problem: the lexicon for nutrient names must enumerate German, English, French, Dutch, and other language forms of every nutrient the pipeline expects to see, and missing any one of them means the corresponding tokens fall through the rules. \textbf{Maintenance cost} is the third limitation: as the lexicon grows, the developer's mental model of which rules fire on which tokens grows in complexity, and adding new rules without disrupting existing behaviour requires careful regression testing on a representative held-out subset.

\subsection{Learned NER as a Classifier Alternative}\label{sec:bg-class-ner}

The second classifier family is \textbf{zero-shot learned named-entity recognition}. A neural model trained on labelled NER data from many domains is presented at inference time with a list of natural-language entity-type strings --- for example, \texttt{nutrient name}, \texttt{unit of measurement}, \texttt{numeric quantity} --- and asked to label spans of the input text as belonging to one of those types. The model has not been trained on supplement labels and has not been fine-tuned for the schema of this thesis; the \emph{task} of mapping spans to user-supplied types is what the model has learned to perform, and the supplement-label schema is one instance of that task.

Modern zero-shot NER models are typically built on transformer encoders. The encoder consumes the input tokens and produces a contextual representation for each one --- a vector that captures the token's meaning in the context of the surrounding tokens, rather than in isolation. Transformer-based contextual representations are what make zero-shot generalisation possible: a token whose surface form was never seen during training can still be classified correctly if the surrounding context resembles contexts the model encountered. Pre-trained transformer encoders such as BERT \cite{devlin2019bert} and its multilingual descendants are the foundation of this capability, and the representation they produce is the input substrate on which span-classification heads operate.

The canonical zero-shot NER system relevant to this thesis is \textbf{GLiNER} \cite{zaratiana2023gliner}, a compact generalist NER model that uses a bidirectional transformer to jointly represent input text and a list of natural-language entity-type prompts supplied at inference time. The model performs parallel entity extraction: candidate spans in the input text are scored against each prompted entity type, and a span is committed as belonging to a type when its score exceeds a threshold and the type is the argmax over candidate types. Because the entity types are specified as natural-language strings rather than as a fixed schema fitted on training data, GLiNER can be redirected to a new schema by changing the prompts rather than by retraining. The thesis evaluates GLiNER as one candidate replacement for the rule-based classifier of \S\ref{sec:bg-class-rules}.

Zero-shot learned NER's strengths are the converses of rule-based classification's limitations. \textbf{Lexicon coverage is no longer a bottleneck}: a nutrient name not enumerated in any lexicon can still be classified correctly if its surrounding context resembles other nutrient names the underlying transformer has been exposed to during pre-training and NER fine-tuning. \textbf{Schema flexibility} is also stronger: a new entity type --- say, an allergen warning that the rule-based classifier was never written to handle --- can be added to GLiNER's prompt without code changes. The costs are the converse of the rule-based virtues: classification decisions are no longer traceable to a specific rule, and the model's behaviour on a new image is harder to predict than the rule-based classifier's. The maintenance burden shifts from lexicon engineering to prompt engineering and threshold tuning. A further cost specific to a graph-based pipeline is that GLiNER, like most learned NER models, operates on a serialised text representation rather than on the OCR token stream with bounding boxes. Predicted spans must therefore be mapped back to the OCR tokens that produced them so that the downstream graph stage can continue to reason about geometry --- a non-trivial alignment step that adds engineering surface area and introduces additional failure modes when the serialisation reorders tokens or merges adjacent spans across what were originally separate OCR regions.

\subsection{Embedding-Based Classification (Brief Overview)}\label{sec:bg-class-embed}

The third classifier family is \textbf{embedding-based prototype matching}. Each token, phrase, or normalised string from the input is mapped to a fixed-dimensional dense vector by a pre-trained multilingual sentence encoder. Each semantic category is represented not by a single prototype but by a \emph{prototype pool} --- a set of natural-language exemplars covering the surface forms the category contains: known nutrient names across the languages of interest for the NUTRIENT category, unit symbols and their variants for the UNIT category, context phrases such as \texttt{per 100g} and their multilingual equivalents for the CONTEXT category. Each member of the prototype pool is mapped to a dense vector by the same encoder, and the category's representation in vector space is the union of these vectors rather than a single prototype centroid. Classification is performed by cosine similarity: a token is assigned to the category some member of whose prototype pool it is closest to, provided the similarity exceeds a threshold and the closest pool member is more similar than any pool member of any other category by at least a margin.

The approach is zero-shot in a different sense than rule-based classification or zero-shot NER. The encoder has been pre-trained on diverse multilingual text but has not been trained on supplement labels or on the specific schema of this thesis; the prototype pools are written by the developer rather than learned; and the cosine-similarity decision rule is non-parametric with respect to fitted model weights --- the threshold and margin are tuned but not fitted by gradient descent. The zero-shot capacity comes from the encoder's general-purpose semantic representations, which can place an unseen nutrient name closer to a member of the NUTRIENT pool than to any member of any other pool because the encoder has learned --- during pre-training on general text --- the lexical and semantic regularities the prototype embeddings exploit. The technical details of this approach, the threshold and margin parameters that govern the decision rule, and the trade-offs that its evaluation in this thesis revealed are the subject of \S\ref{sec:bg-embeddings}.

\subsection{Three Faces of Zero-Shot Classification}\label{sec:bg-class-zeroshot}

The three classifier families introduced in \S\ref{sec:bg-class-rules}, \S\ref{sec:bg-class-ner}, and \S\ref{sec:bg-class-embed} are all called zero-shot in the document-IE and NLP literature, but the word means three different things across them. The differences matter for understanding the design space, because each family makes a different bet about what kind of prior knowledge is most useful for this particular task.

The \textbf{rule-based classifier} is zero-shot because no parameters are fitted from labelled data. The prior knowledge it exploits is whatever the developer has written into the lexicons, regular expressions, and decision rules. This prior is \textbf{explicit} in the strongest sense: every classification decision is the consequence of a rule that exists in source code and can be inspected, modified, or reasoned about by any reader of the codebase. The rule-based classifier's capacity is bounded by the breadth of the developer's domain knowledge and by the patience of the developer to enumerate it.

\textbf{Zero-shot learned NER} is zero-shot in a different sense: it has been trained on labelled NER data from many domains, but not on the schema of this thesis. The prior knowledge it exploits is \textbf{transferred} --- the model's parameters encode statistical regularities of the entity-recognition task that are presumed to generalise across schemas, supplemented at inference time by the natural-language entity-type strings the user supplies. The classifier's capacity is bounded by how well its training distribution covers the inferential regularities the supplement-label task requires.

\textbf{Embedding-based prototype matching} is zero-shot in a third sense: a multilingual sentence encoder has been pre-trained on diverse general-purpose text, and its embeddings are used as a similarity space against developer-written prototype pools. The prior knowledge it exploits is \textbf{distributional} --- the encoder has learned lexical and semantic regularities from a large pre-training corpus, and the cosine-similarity decision rule depends on those regularities being present in the embedding space the encoder produces. The classifier's capacity is bounded by what the encoder has learned to represent and by the developer's skill at writing prototypes that exploit those representations.

The three families therefore stake their generalisation on three different sources: hand-coded rules, cross-task statistical transfer, and pre-trained embedding geometry. None of these sources is intrinsically superior; each is appropriate to a different setting, and the empirical question of which one works best for the supplement-label domain is examined in the experimental chapter. What \S\ref{sec:bg-classification} establishes here is that the three approaches are not interchangeable variants of one technique. They embody different commitments about what counts as legitimate prior knowledge for a zero-shot system to bring to a new task, and the choice between them is therefore a methodological commitment as much as an engineering one.

% =========================================================================
\section{Graph Representations of Documents}\label{sec:bg-graphs}
% =========================================================================

\subsection{Graphs as a Document Representation}\label{sec:bg-graphs-intro}

The pipeline of this thesis solves two problems: extraction and association. Extraction asks what each token means in isolation --- is this token a nutrient name, a numeric value, a unit, or something else? Association asks how the tokens that have been extracted relate to one another --- which nutrient does this quantity belong to, which unit accompanies which value, which header governs which column of data. The two problems have different shapes: extraction is \emph{classificatory}, in that each token or token span is assigned a semantic category; association is \emph{relational}, in that the meaning of any single relationship depends on the surrounding context of other relationships. The natural data structure for relational reasoning is the \textbf{graph}.

A graph in the document-IE setting consists of \textbf{nodes}, which are the OCR tokens after classification has assigned them a semantic role, and \textbf{edges}, which encode relationships between tokens. The pipeline of this thesis uses several edge types --- for example, \texttt{SAME\_ROW} between tokens in the same logical row, \texttt{SAME\_COL} between tokens in the same column, and \texttt{CONTEXT\_SCOPE} between a header and the data tokens it governs --- and the full set is enumerated in the methodology chapter. The graph is therefore not merely a description of the document's appearance but an explicit representation of its inferred structural relationships, with the geometric primitives of \S\ref{sec:bg-layout} providing the inputs from which most of the edges are constructed.

Recasting the association problem as a graph problem is what allows the pipeline to use the same machinery across very different layouts. A flat table, a multi-column panel, and a paragraph-style label all produce different distributions of node positions, but the graph operations that consume the resulting node-and-edge structures --- pruning, matching, propagating context --- are the same regardless of where the nodes came from. The remainder of \S\ref{sec:bg-graphs} develops the specific operations these graphs support.

\subsection{Edge Types and Soft Adjacency}\label{sec:bg-graphs-edges}

Edges in a document graph carry meaning beyond mere connection between two nodes. Two tokens that share a \texttt{SAME\_ROW} edge encode a different relationship than two tokens that share a \texttt{SAME\_COL} edge or a \texttt{CONTEXT\_SCOPE} edge, even when the underlying token pair is the same. The graph constructor therefore labels each edge with a \textbf{type} drawn from a fixed set, and the association stage that consumes the graph is sensitive to type: pruning operates differently on row edges than on column edges, and matching operates differently on context-scope edges than on either. Edge types are how a single graph carries multiple kinds of structural information without conflating them.

Each edge also carries a \textbf{weight} --- a continuous scalar between 0 and 1 --- that quantifies how strongly the relationship encoded by the edge applies to the specific token pair. The use of weighted rather than binary edges is a methodological choice related to the soft-adjacency idea used in graph-based document IE \cite{yu2020pick}, in which adjacency is treated as a continuous-valued relationship between tokens rather than a binary one. PICK realises soft adjacency by \emph{learning} the adjacency weights end-to-end from text, visual, and layout features. The pipeline of this thesis adapts the same idea to a deterministic, zero-shot setting: the weights are computed by closed-form functions of geometry rather than learned, but the underlying commitment --- that the strength of a pairwise relationship varies continuously rather than binary --- is preserved. The contrast that motivates this choice is between two ways of deciding whether two tokens are in the same row.

The simpler approach, \textbf{hard adjacency}, applies a binary distance threshold. Two tokens are in the same row if and only if the absolute vertical distance between their centroids is below some fixed value, say 25 pixels; otherwise they are not. Hard adjacency has the advantage of simplicity and the disadvantage of brittleness. A label printed in a slightly larger font has rows separated by 30 pixels and breaks the threshold; a label printed in a slightly smaller font has rows separated by 12 pixels and chains adjacent rows together. Worse, on a label whose row spacing is genuinely heterogeneous --- header rows tall, data rows short, sub-nutrient rows tighter still --- no single threshold value can be correct for all of them.

A weighted formulation instead assigns each candidate edge a weight that decays continuously with distance. Two tokens whose vertical centroid difference is small carry a high-weight \texttt{SAME\_ROW} edge; two tokens whose difference is moderate carry a low-weight edge; two tokens whose difference is large carry no edge at all. The weight is computed by a deterministic function --- typically a function that returns 1 at zero distance, decays smoothly as distance grows, and reaches 0 at some cutoff --- that is calibrated against a \emph{local scale} estimated from the surrounding neighbourhood, such as median token height or characteristic row spacing. The local-scale calibration is what allows the weight function to behave consistently across labels with different font sizes and different row spacings.

The pipeline of this thesis uses weighted edges of this kind for \textbf{spatial-distance edges} --- \texttt{SAME\_ROW}, \texttt{SAME\_COL}, and others where the edge's existence and strength are functions of geometric proximity --- because hard distance thresholds are exactly what break under the geometric distortions of \S\ref{sec:bg-layout-distortions}. Other edge types, including \texttt{CONTEXT\_SCOPE}, are \textbf{structurally inferred}: the edge endpoints are spatially grounded --- a header is physically above the data tokens it governs --- but the decision to create the edge is made by the column-role inference of the methodology stage, not by a distance computation. The weight of a structurally inferred edge either carries a fixed nominal value or reflects the confidence of the column-role inference itself. Weighted spatial adjacency and structurally inferred edges therefore coexist in the same graph, each appropriate to a different category of relationship.

\subsection{Tail-Collision Avoidance}\label{sec:bg-graphs-tca}

A weighted graph constructed from a dense supplement label routinely contains contradictions: edges that, if all retained, would violate the one-to-one structural constraints the document actually obeys. The most common contradiction is the \textbf{tail collision}: two nutrient nodes both connect, through high-weight \texttt{SAME\_ROW} edges, to the same quantity node. The graph asserts that the quantity belongs to nutrient A \emph{and} that it belongs to nutrient B. Both edges are individually plausible --- both nutrients are spatially close enough to the quantity for the soft-adjacency function to assign each a non-trivial weight --- but only one of them can be correct on a label that reports each nutrient with one quantity per context. The graph is internally inconsistent, and a downstream matching stage that consumes the graph naively will commit one of the two errors: assigning the quantity to the wrong nutrient, or splitting it ambiguously between both.

The technique that resolves this contradiction at the graph level, before matching, is \textbf{Tail-Collision Avoidance (TCA)}, introduced by SPADE \cite{hwang2021spade} as a deterministic edge-resolution step in spatial-dependency parsing of semi-structured documents. SPADE motivates TCA from the observation that, in semi-structured documents, a node almost always has a single incoming edge per relation --- with tabular structures being a notable exception. The original SPADE algorithm is iterative: at each tail node carrying multiple incoming edges, the highest-probability incoming edge is retained, the remaining edges are trimmed, the trimmed head nodes are redirected to their next-most-probable tails, and the process repeats until the graph becomes self-consistent or an iteration limit is reached. The pipeline of this thesis adopts a simplified, single-pass adaptation suited to its zero-shot setting: when two edges of the same type share a tail node, only the higher-weight edge is retained, and the lower-weight one is dropped, with no redirection step. This is sufficient for supplement-label tables, where the dense local geometry means that the trimmed head nodes are typically already covered by other high-weight edges from the surrounding row structure.

TCA is run after the soft-adjacency edges are constructed and before the bipartite matching stage of \S\ref{sec:bg-graphs-matching}. The result of the simplified procedure is a graph in which each target node has at most one retained incoming edge of the relevant type. TCA is also type-sensitive: it operates on \texttt{SAME\_ROW} edges and \texttt{SAME\_COL} edges separately, because the one-to-one constraint applies within each type rather than across them. A nutrient may legitimately have a \texttt{SAME\_ROW} edge to one quantity and a \texttt{SAME\_COL} edge to a different one --- that is how multi-context labels are structured --- and TCA must not collapse such structures. The simplified TCA adaptation is purely a function of edge weights and edge types, and is therefore agnostic to whether those weights came from a learned graph module or from a deterministic geometric function.

\subsection{Bipartite Matching, with Rank-Under-Context as Fallback}\label{sec:bg-graphs-matching}

After tail-collision avoidance has resolved the within-type contradictions of \S\ref{sec:bg-graphs-tca}, the graph still contains a question that no edge-pruning step can answer alone: which of the surviving high-weight edges constitute the final set of associations that become the output tuples? Different field associations have different structural shapes. Nutrient-to-quantity association is one-to-one \emph{within a given context column} --- each nutrient has at most one quantity in the per-100g column and at most one in the per-serving column. Quantity-to-unit association takes one of two structural shapes depending on the label's typography. When each quantity is printed adjacent to its own unit (\texttt{12 g}, \texttt{8 g}), the association is one-to-one within a row. When the unit is given once in a column header and inherited by every quantity beneath, the association is one-to-many propagation rather than one-to-one matching, structurally identical to the header-to-data context case. And header-to-data context association is itself one-to-many --- a single \texttt{per\_100g} header governs every quantity token in its column. The matching machinery must therefore recognise the structural shape of each association before choosing how to resolve it.

For the one-to-one associations, the natural formal framework is \textbf{bipartite matching}: two sets of nodes connected by weighted edges of a given type, where the task is to choose a subset of edges such that no node is incident to more than one chosen edge and the total weight of the subset is as large as possible. The canonical optimum-finding algorithm for this problem is the \textbf{Hungarian algorithm} \cite{kuhn1955hungarian}, which finds the maximum-weight matching in polynomial time. The pipeline of this thesis uses a simpler \textbf{greedy bipartite-matching heuristic}, faithful to the matching algorithm of \S\ref{sec:bg-evaluation-tuplef1}: edges are sorted by weight from highest to lowest, and each edge is committed to the matching if and only if neither of its endpoints has already been committed to an earlier edge. The procedure terminates when no remaining edge has positive weight or when one of the two node sets is exhausted. The greedy heuristic is suboptimal in pathological cases --- it can commit early to a high-weight edge that prevents a globally better matching --- and the gap to the Hungarian optimum is treated as an empirical trade-off rather than a proven small quantity. The simplicity of the greedy procedure is preserved consistently across the pipeline and the evaluator, both of which use the same algorithm.

For the one-to-many associations --- header-to-data context propagation, and the header-inherited unit case described above --- bipartite matching is not the right framework. Headers govern all the data tokens in their column simultaneously, so the matching is ``every member of column $k$ receives the context (or unit) of column $k$'s header'' rather than ``each data token chooses a single edge to commit''. The pipeline handles these cases through structural propagation rather than through edge-by-edge matching: once column-role inference has identified each column's header, the header's value is propagated to every data token in the column as a property, not as a chosen edge.

Bipartite matching on geometric edges works well when the underlying geometry is reliable. On a flat label with clean OCR, the highest-weight \texttt{SAME\_ROW} edge from each nutrient leads to its correct quantity, and greedy matching commits the correct pairs in order. The matching breaks when the geometry is unreliable: severely curved labels where the row inducer has under-merged or over-merged, OCR noise that has scrambled the centroid positions of some tokens, or layouts where the per-serving column drifts so far that the \texttt{SAME\_ROW} edges link tokens that are visually adjacent but logically belong to different rows. In all of these cases, the geometric edges either cannot be constructed or carry weights that mislead the matching.

The pipeline therefore retains a \textbf{structural fallback} for cases where geometric matching fails: \textbf{rank-under-context}. The principle is to match tokens not by their spatial position but by their ordinal rank within their column, conditioned on the shared column context. Concretely, the nutrients in the nutrient column are sorted top-to-bottom and given ranks $1, 2, 3, \ldots$; the quantities in the per-100g column are sorted similarly and given the same ranks; the quantities in the per-serving column likewise. The rank-1 nutrient pairs with the rank-1 per-100g quantity and the rank-1 per-serving quantity; the rank-2 nutrient with the rank-2 quantities; and so on. Rank-under-context recovers tuples in cases where the per-100g and per-serving columns have drifted so far apart geometrically that no \texttt{SAME\_ROW} edge can usefully cross the gap, where geometric matching alone would fail.

Rank-under-context is not safe to apply unconditionally. The rule depends on the column rank-lists being mutually compatible: each column must contain the same number of data entries as the others, in the same vertical order, with no row-skip caused by a missing OCR token or a sub-row absorbed into the wrong column. When these preconditions fail, rank-1 in one column and rank-1 in another no longer correspond to the same logical row, and rank-based matching produces tuples that look correct on per-column counts but are misaligned on the actual label. The pipeline therefore engages rank-under-context only when two conditions hold: the column-role inference of the methodology stage indicates that each column is well-defined and well-populated, and the rank lists across the relevant columns have compatible lengths. When either condition fails, the affected tokens are left unmatched rather than matched incorrectly, on the principle that a missing tuple is more recoverable than a wrong one.

The result of these design choices is a hybrid matcher: the greedy bipartite path handles the typical case where geometry is reliable, the structural propagation path handles the one-to-many associations that bipartite matching cannot cover, and the rank-under-context fallback handles the cases where geometric matching has failed but column structure is still recoverable. The three modes complement rather than compete, and each is engaged only when its preconditions are met.

\subsection{Deterministic versus Learned Graph Reasoning}\label{sec:bg-graphs-detlearn}

The graph-reasoning primitives of \S\ref{sec:bg-graphs-intro} through \S\ref{sec:bg-graphs-matching} --- typed edges with continuous weights, tail-collision avoidance, bipartite matching, structural propagation, rank-under-context --- are situated within a body of work on graph-based key-information extraction in the document-understanding literature, and the most directly relevant prior systems are PICK \cite{yu2020pick} and SPADE \cite{hwang2021spade}. PICK formulates key-information extraction by combining graph learning and graph convolution over text segments, with edge weights produced from textual, visual, and layout features and learned end-to-end from labelled documents. SPADE formulates semi-structured document IE as spatial dependency parsing over OCR tokens and their 2D coordinates, predicting head--tail dependency edges and including tail-collision avoidance as part of the graph-decoding step. Both systems are supervised: they require labelled examples of the target schema and learn the relevant graph machinery from those examples. The pipeline of this thesis differs from both in that it constructs the graph deterministically from geometric and lexical primitives, without any labelled training data, and combines these primitives with the structural-propagation and rank-under-context mechanisms of \S\ref{sec:bg-graphs-matching} in a configuration that is specific to this thesis's design.

The trade-off between learned and deterministic graph reasoning is the same one that runs through every stage of the pipeline. Supervised graph learning can achieve stronger performance on the datasets it is trained for: a graph module trained end-to-end on labelled receipts or invoices learns weights that capture domain-specific patterns that are difficult to encode manually. The cost is that supervised systems may not generalise reliably across domains without adaptation or retraining, and that they require labelled data the supplement-label domain does not have. The thesis adopts the deterministic alternative: it borrows the \emph{architectural ideas} --- soft adjacency from PICK, tail-collision avoidance from SPADE, the broader principle that document IE benefits from typed-edge graph reasoning --- and re-implements them as functions of geometry and semantic role rather than as learned parameters. The architectural commitments survive the adaptation; the supervised-learning machinery does not.

This adaptation is not a degradation of PICK or SPADE; it is a different point in the design space. The architectural ideas of soft adjacency and tail-collision avoidance are independent of \emph{how} edge weights are produced, and the same graph operations that consume a learned adjacency matrix can also consume a deterministic one, provided the scores are comparable within the pruning or matching step. What changes is the calibration: a learned weight reflects the network's training experience, while a deterministic weight reflects explicit assumptions about the geometric regularities of the input domain. The two converge when those assumptions sufficiently capture the relevant geometry; they diverge when the input contains regularities the assumptions have not anticipated. The empirical question of which regime is preferable for the supplement-label domain --- where the input is geometrically constrained but variable, and where labelled data is scarce --- is examined in the experimental chapter.

% =========================================================================
\section{Embeddings and Semantic Similarity}\label{sec:bg-embeddings}
% =========================================================================

\subsection{Embeddings: From Word Vectors to Multilingual Sentence Encoders}\label{sec:bg-embed-history}

An \textbf{embedding} is a fixed-dimensional vector representation of a textual unit --- a word, a phrase, a sentence, or a longer passage --- produced by a model that has been trained such that useful semantic or distributional relationships are reflected in the vector space. The fundamental utility of embeddings is that the geometry of the vector space carries semantic information: two strings that mean similar things often land at similar positions, and the cosine of the angle between their vectors often becomes a usable proxy for how similar their meanings are. This is what allows downstream stages to make similarity-based decisions without enumerating explicit rules --- the model's training has effectively compiled enormous amounts of distributional information about natural language into geometric relationships.

Early embedding models operated at the \textbf{word level}. Word2Vec \cite{mikolov2013word2vec} and GloVe \cite{pennington2014glove} learned a single fixed vector for each word in a vocabulary, by either predicting words from their contexts or counting co-occurrences in a large corpus. These models established the basic property that a vector space could be made to reflect linguistic regularities, and the famous analogy $\textit{king} - \textit{man} + \textit{woman} \approx \textit{queen}$ is the canonical illustration that semantic relationships transfer to vector arithmetic. The limitation of word-level embeddings is the per-word fixed vector: a word with multiple senses receives a single representation that conflates them, and a word never seen during training has no embedding at all.

The dominant embedding paradigm of the modern era is \textbf{contextual}. Transformer-based models such as BERT \cite{devlin2019bert} produce a different vector for each token \emph{in the context} of the surrounding tokens --- the same word in different contexts produces different embeddings, and an unseen word is represented by composing its subword pieces, each of which the model has been exposed to. Contextual embeddings inherit the geometric semantics of word embeddings but reduce their limitations: polysemy is handled implicitly, and the open-vocabulary problem is mitigated by the subword tokeniser. The cost is that a token's embedding now depends on its surrounding text, so embeddings cannot be precomputed and looked up; they must be computed by running the model on the input each time.

A further development relevant to this thesis is the \textbf{sentence encoder} --- a model that produces a single fixed-dimensional vector for an entire sentence or short phrase, rather than one vector per token. Sentence-BERT \cite{reimers2019sentencebert} introduced the standard recipe: take a pre-trained transformer encoder, pool the per-token outputs into a single vector, and fine-tune the result on sentence-pair similarity tasks so that the pooled vectors are aligned with semantic similarity rather than with the encoder's original masked-language-modelling objective. The sentence encoder is the natural unit of representation for prototype-matching classification, where each prototype is a short natural-language phrase and each candidate is a token, span, or normalised string.

The final development this thesis builds on is the \textbf{multilingual sentence encoder}. A multilingual encoder is one that has been trained on text in many languages such that semantically equivalent strings in different languages land at nearby positions in the embedding space. This property is essential for supplement labels, where the same nutrient may appear in German on one product, English on another, and Italian on a third, and a classifier whose generalisation depends on cross-lingual semantic alignment requires an encoder whose embedding space respects that alignment. The specific multilingual sentence encoder this thesis uses, BGE-M3, is the subject of the next subsection.

\subsection{BGE-M3 and Cosine Similarity}\label{sec:bg-embed-bgem3}

The multilingual sentence encoder used in this thesis is \textbf{BGE-M3} \cite{chen2024bgem3}, a publicly released model whose name reflects three design properties advertised by its authors: \emph{multi-functionality} (it produces three kinds of representation --- dense, sparse, and multi-vector --- though the pipeline of this thesis uses only the dense representation), \emph{multi-linguality} (it has been trained on text in over a hundred languages), and \emph{multi-granularity} (it accepts inputs ranging from individual tokens to passages of several thousand tokens). For each input string, the model produces a single dense vector of fixed dimension that the pipeline uses as the input string's embedding.

BGE-M3 is the result of a self-knowledge-distillation training scheme that combines retrieval supervision in many languages with a unified loss across the three representation types. The practical consequence for this thesis is that the encoder produces dense vectors that are aligned across languages: a German string and its English equivalent tend to land near one another in the vector space, and the cosine similarity between them tends to be higher than between unrelated strings, which is the property the embedding-based classifier depends on. The prototype pools introduced in \S\ref{sec:bg-class-embed} contain exemplars in several languages, and a candidate token must be matchable against any of them regardless of the language it itself appears in.

The metric used to compare two embeddings produced by BGE-M3 in its dense mode (which, as noted above, is the only representation type the pipeline of this thesis uses) is \textbf{cosine similarity}. Given two vectors $\mathbf{u}$ and $\mathbf{v}$, the cosine similarity between them is defined as
\begin{equation*}
  \cos(\mathbf{u}, \mathbf{v}) \;=\; \frac{\mathbf{u}^{\top}\mathbf{v}}{\|\mathbf{u}\|\,\|\mathbf{v}\|},
\end{equation*}
the inner product of the vectors normalised by their lengths. The result is a scalar in the interval $[-1, 1]$, where $1$ indicates identical direction in vector space, $0$ indicates orthogonality, and $-1$ indicates opposite direction.

Cosine similarity is preferred over Euclidean distance for embedding comparison for two reasons. First, sentence encoders such as BGE-M3 are trained with similarity objectives that operate on direction in vector space rather than on magnitude, so the \emph{angle} between two vectors is the quantity the model has been trained to make meaningful, while the \emph{length} of either vector is a less reliable signal. Second, cosine similarity is invariant to vector scaling: multiplying $\mathbf{u}$ by any positive constant leaves $\cos(\mathbf{u}, \mathbf{v})$ unchanged, which means the comparison is robust to the per-input length variation that arises from inputs of different lengths or token counts. Euclidean distance possesses neither property, and embeddings compared by Euclidean distance often produce counter-intuitive rankings in which a longer vector dominates over a semantically closer but shorter one.

\subsection{Prototype Matching, Thresholds, and Margins}\label{sec:bg-embed-prototype}

The embedding-based classifier's role within the broader classifier of \S\ref{sec:bg-classification} is restricted to \textbf{nutrient detection}: deciding, for each candidate string in a position where a nutrient could plausibly appear, whether the string represents a nutrient or not. The output is binary --- either the candidate is committed as a NUTRIENT, or it is rejected and routed to UNKNOWN. The classifier does not identify \emph{which} nutrient the candidate represents; that decision, when needed downstream, is performed by lexicon lookup against the canonical nutrient names. Other classification roles --- UNIT, CONTEXT, QUANTITY, NOISE --- remain rule-based even when the embedding classifier replaces the nutrient-detection step.

The classifier maintains a single prototype pool $\mathcal{P}$, written by the developer, containing natural-language exemplars of nutrients across the languages of interest, including common abbreviations and chemical synonyms. At classifier startup, every member of $\mathcal{P}$ is embedded by BGE-M3 once and the resulting vectors are cached, so that classification at inference time requires only the embedding of the candidate.

For a candidate string with embedding $\mathbf{u}$, the classifier computes the score
\begin{equation*}
  s(\mathbf{u}) \;=\; \max_{p \in \mathcal{P}} \cos(\mathbf{u}, \mathbf{u}_p),
\end{equation*}
the cosine similarity to the \emph{closest} prototype in the pool. The classifier commits the candidate as NUTRIENT if and only if $s(\mathbf{u}) \geq \tau$, where $\tau$ is a threshold parameter. If $s(\mathbf{u}) < \tau$, the candidate is classified as UNKNOWN.

The threshold $\tau$ is the only parameter governing the decision. A high $\tau$ rejects candidates whose closest nutrient prototype is too dissimilar --- typical of OCR garble, fragments of non-nutritional prose, and out-of-distribution inputs the encoder cannot interpret --- and is the principal defence against false-positive nutrient detections. A low $\tau$ admits more candidates as nutrients, recovering some that lie outside the prototype pool's coverage at the cost of false positives. The trade-off is one-dimensional, which is what makes the threshold tuneable through a one-dimensional sweep rather than the two-dimensional grid that would be required for a margin-based discriminator.

The threshold is set by \textbf{grid search on a held-out validation subset} of the labelled data. A range of $\tau$ values is enumerated, the classifier is evaluated on the validation subset under each value, and the value that maximises 4F~F1 (or 3F~F1, depending on the experiment's objective) is selected as the operating point. The held-out validation subset is disjoint from the evaluation/test set used for final reporting in the experimental chapter, so threshold selection does not leak information from the evaluation data into the reported metrics. This procedure is the only point in the embedding-based classifier where labelled data is consulted: BGE-M3 was pre-trained without supplement-label data, the prototype pool is written by the developer, and the cosine-similarity decision rule is training-free with respect to fitted model weights. The threshold selection is, however, an explicit form of hyperparameter tuning on labelled data, which means the embedding-based classifier sits at a slightly weaker zero-shot point in the design space than the rule-based classifier of \S\ref{sec:bg-class-rules} --- a methodological caveat acknowledged in \S\ref{sec:bg-problem}'s zero-shot commitment paragraph.

The simplicity of a one-dimensional threshold sweep is what makes the embedding-based classifier amenable to systematic ablation. The experimental chapter presents threshold sweeps that hold all other components of the pipeline fixed and vary only $\tau$, allowing the resulting performance curve to be plotted explicitly. The shape of this curve --- where the optimum lies, how broad the high-performing region is, how steeply performance falls off on either side --- is itself an empirical signal about how well the encoder's embedding space distinguishes nutrient inputs from non-nutrient inputs.

\subsection{Failure Modes of Prototype Matching}\label{sec:bg-embed-failure}

The embedding-based classifier's performance depends on properties of the encoder's embedding space that are not under the developer's control. When those properties hold for a particular candidate string, the binary nutrient/non-nutrient decision is reliable; when they fail, the classification can be wrong in ways that no amount of threshold tuning can fully correct. This subsection names the principal failure modes that arise in practice on supplement-label data, both as a calibration of expectations for the experimental chapter and as context for the methodological choices \S\ref{sec:bg-classification} has already laid out.

The first failure mode is \textbf{out-of-distribution candidates rejected as UNKNOWN}. A candidate string that consists of OCR garble, partial words, or fragments of non-nutritional prose typically produces an embedding that is not close to any prototype in the pool. The maximum cosine similarity is below the threshold $\tau$, the decision rule fails, and the candidate is routed to UNKNOWN. This is the \emph{intended} behaviour: the threshold is a defence against forced commitments on inputs the encoder cannot interpret. But it is also a failure mode in the sense that some out-of-distribution candidates carry latent nutrient identity that a more aggressive classifier might recover. A token like \texttt{vit.B12*} after artefact stripping might genuinely refer to vitamin B12, but if the encoder embeds it far from the nutrient prototypes --- because the truncated form and the asterisk artefact disrupt the encoder's expected input distribution --- the classifier rejects it. Lowering $\tau$ to recover such cases comes at the cost of admitting genuinely non-nutrient inputs as false positives.

The second failure mode is \textbf{non-nutrient candidates falsely admitted as NUTRIENT}. Because the binary decision partitions the embedding space into a nutrient region (above the threshold) and a non-nutrient region (below it), any non-nutrient candidate that the encoder happens to embed close to a nutrient prototype will be misclassified as a nutrient. The most common cases on supplement labels involve tokens that lie in the same semantic neighbourhood as nutrients without being nutrients themselves: chemical names of non-nutrient compounds, regulatory or marketing claims that mention nutrients indirectly (``supports vitamin C absorption''), and prose fragments that cite a nutrient in a non-extractive context. Tightening $\tau$ to reject these false positives risks rejecting genuine nutrients whose embeddings are also moderately close to non-nutrient regions of the space --- the trade-off is the threshold's central tension.

The third failure mode is \textbf{multilingual drift}. BGE-M3's cross-lingual alignment is a tendency rather than a guarantee. The drift can manifest in either direction. A genuine nutrient term in a language sparsely represented in the prototype pool may produce an embedding that is closer to a non-nutrient region than to the pool's nearest prototype, leading to a false rejection. Conversely, a non-nutrient term whose embedding the encoder has placed close to nutrient-region prototypes for cross-lingual reasons --- perhaps because the term shares lexical or distributional patterns with nutrient names in another language --- may be falsely admitted as a nutrient. This failure mode is hardest to diagnose and is the one most directly tied to the encoder's training distribution.

These three failure modes do not exhaust the ways binary nutrient prototype matching can fail, but they account for the kinds of misclassification that arise routinely on the supplement-label domain. The methodological response is twofold. First, the threshold sweep of \S\ref{sec:bg-embed-prototype} maps the operating-point range and identifies the value that best balances precision and recall in aggregate, accepting that no single threshold eliminates the failure modes individually. Second, the rule-based classifier of \S\ref{sec:bg-class-rules} remains the production choice precisely because its failure modes are different in kind --- lexicon coverage gaps and surface-form variation rather than embedding-space pathologies --- and the experimental chapter compares the two failure-mode profiles empirically to inform which approach is better suited to the supplement-label task.

% =========================================================================
\section{Evaluation}\label{sec:bg-evaluation}
% =========================================================================

\subsection{Precision, Recall, and F1}\label{sec:bg-evaluation-prf1}

Every metric reported in this thesis is built from three quantities: precision, recall, and F1. Let $\mathcal{P}$ be the set of predicted tuples produced by the pipeline on a test image, and let $\mathcal{G}$ be the set of ground-truth tuples annotated for that image. A \emph{matching} $M \subseteq \mathcal{P} \times \mathcal{G}$ pairs predictions with ground-truth tuples under a chosen \emph{match predicate} --- a binary function that decides, for any $(p, g)$ pair, whether $p$ counts as a correct extraction of $g$. From the matched count $|M|$, the standard definitions follow:
\begin{equation*}
  P = \frac{|M|}{|\mathcal{P}|}, \quad R = \frac{|M|}{|\mathcal{G}|}, \quad F_1 = \frac{2\,P\,R}{P+R}.
\end{equation*}
Precision answers ``of all the tuples the pipeline emitted, what fraction were correct?''; recall answers ``of all the tuples the ground truth contains, what fraction did the pipeline find?''; F1 is the harmonic mean and penalises any imbalance between the two. What distinguishes one evaluation regime from another is not the formulas --- these are fixed across the field --- but the match predicate. The remainder of this section makes the predicates of this thesis explicit.

\subsection{Token-, Span-, and Tuple-Level Predicates}\label{sec:bg-evaluation-predicates}

Three families of match predicates are common in the document-IE literature, and they differ in how strictly they require a prediction and a ground-truth item to agree. A \emph{token-level} predicate counts individual labelled tokens: a token is correct when its predicted label matches its ground-truth label, regardless of which entity that token belongs to or whether the surrounding entity span is correctly bounded. A \emph{span-level} predicate counts correctly bounded entity spans: a span is correct when its boundaries and its label both match the ground truth, but the span is evaluated in isolation from any other span in the document. A \emph{tuple-level} predicate counts structured records: a tuple is correct when \emph{all} fields of the record match the ground truth simultaneously, and the predicted fields actually correspond to the same underlying fact on the page rather than being extracted from unrelated parts of the document. The three predicates form a strict hierarchy of difficulty --- every span-level error is also a token-level error, and every tuple-level error is also at least one span-level error --- but the converse does not hold, and gains at one level do not automatically translate to gains at another.

This thesis reports both token-level and tuple-level F1, but the headline metric is the tuple-level score. The reason for keeping both is that they answer different questions. Token-level F1 measures the classifier in isolation: it tells you whether the right tokens are being labelled with the right semantic categories, independently of how the pipeline later assembles them. Tuple-level F1 measures the pipeline as a whole: it tells you whether the final structured output is correct, which is the question a downstream consumer of the pipeline actually cares about. The two metrics can move in opposite directions. A more aggressive classifier can improve token-level F1 by labelling more tokens as nutrients, but if those extra tokens are noise that the graph constructor cannot place coherently, tuple-level F1 falls. Conversely, a stricter classifier that drops borderline tokens can hurt token-level F1 but produce cleaner tuples downstream. The thesis reports both so that any claim of the form ``stage X improved'' is anchored to the level at which the improvement was measured, and so that the experimental chapter can distinguish a real end-to-end gain from a localised improvement that did not survive to the output.

\subsection{The 4-Field Tuple F1 of This Thesis}\label{sec:bg-evaluation-tuplef1}

The headline metric of this thesis is the F1 of the 4-field tuple. A predicted tuple takes the form $\langle \text{nutrient},\,\text{quantity},\,\text{unit},\,\text{context} \rangle$, and matches a ground-truth tuple if and only if all four fields agree under a normalised comparison. Three-field F1 (3F~F1) --- which drops the context field --- is reported alongside as a secondary metric, both because the context field is the field most sensitive to OCR-driven column-header errors and because the contrast between 3F~F1 and 4F~F1 quantifies, image by image, how much performance is being lost specifically to context-attribution failures.

The match predicate operates field-wise. The nutrient comparison applies a normalisation that lowercases, strips diacritics, folds German umlauts using the standard transliteration convention ({\ss}~$\rightarrow$~ss, {\"a}~$\rightarrow$~ae, {\"o}~$\rightarrow$~oe, {\"u}~$\rightarrow$~ue), removes parenthetical qualifiers, and discards trailing punctuation; if the strings then agree, the field matches. The quantity comparison parses both predicted and ground-truth strings into numeric values --- normalising the German decimal comma to a period and stripping the leading \texttt{<} from trace-amount expressions such as \texttt{<0.5} --- and accepts the parsed values as equal only if they are identical. The unit comparison applies a synonym table that unifies common variants ($\mu$g, mcg, ug; kJ/kcal, kcal/kJ; ml, mL). The context comparison normalises whitespace and applies a translation table mapping surface variants (``PRO PORTION'', ``Je 100g'', ``per 100 g'') onto a closed set of three canonical labels: \texttt{per\_100g}, \texttt{per\_serving}, and \texttt{per\_daily\_dose}.

The matching between $\mathcal{P}$ and $\mathcal{G}$ is one-to-one. Naively, a single predicted tuple could match multiple ground-truth tuples (when the label has duplicate nutrients in different contexts, for example), or multiple predictions could match the same ground-truth tuple. Both cases would inflate the matched count. The thesis follows the standard convention from the document-IE literature: an $N \times M$ score matrix is computed by evaluating the predicate on every $(p, g)$ pair, the matrix is then traversed greedily in score-descending order, and at each step the highest-scoring unmatched pair is committed to the matching $M$ and both rows are removed from further consideration. The procedure terminates when no unmatched pair scores above zero. This guarantees that $|M| \leq \min(|\mathcal{P}|,\,|\mathcal{G}|)$ and that the same prediction cannot be credited twice. The score function used to rank pairs in the matrix is the count of fields that agree under the predicates above, with a small unit-agreement tiebreaker (\texttt{unit\_bonus}) for pairs that tie on field count.

\subsection{Normalisation and the Two-Tier Evaluator}\label{sec:bg-evaluation-twotier}

The match predicates of \S\ref{sec:bg-evaluation-tuplef1} are deterministic: they apply fixed lookup tables, fold a small set of diacritics, and decide on string equality. They handle the bulk of the cases that arise in practice, which is why they constitute the \emph{fast tier} of the evaluator. But they do not handle every case in which a predicted tuple and a ground-truth tuple denote the same nutritional fact. Two pairs of strings can describe the same nutrient and yet fail every deterministic rule: a chemical name and its common name (\texttt{Cobalamin} and \texttt{Vitamin B12}); two equivalent multilingual surface forms after umlaut folding has been applied to one but not the other; a compound nutrient name and its conventional shorthand (\texttt{saturated fatty acids} and \texttt{saturates} on UK and EU labels). A deterministic evaluator marks all of these as mismatches and underestimates the pipeline's true accuracy.

The thesis evaluator therefore adopts a two-tier design. Pairs that match deterministically are accepted by the fast tier and never escalated. Pairs that the fast tier rejects but that \emph{plausibly} denote the same fact --- by some heuristic such as token overlap, edit distance, or shared numeric content --- are escalated to the \emph{LLM tier}, in which a language model judge (Qwen-2.5:7b, served locally via Ollama in the reference implementation) is shown both tuples and asked to return a binary equivalence verdict. The LLM tier is bounded both in scope (only fast-tier rejects are escalated) and in cost (the LLM is queried only on borderline cases, not on every pair).

The cost of the LLM tier is reported alongside the headline F1 as a separate metric, \texttt{context\_cost\_pp} --- the percentage-point increase in 4F~F1 attributable to LLM-accepted matches. This is the maximum amount by which the headline F1 could be inflated if the LLM judge were biased toward over-acceptance. Reporting it separately lets the reader see, for every experiment, what the score would be under the fast tier alone and what the LLM tier added. % [PLACEHOLDER: a single concrete example of a pair the fast tier rejected and the LLM tier accepted, drawn from llm_evaluation_details.csv. Until that inspection is done, the example below is illustrative.]
For example, a predicted tuple $\langle$``Vitamin B12'', ``2.5'', ``$\mu$g'', \texttt{per\_daily\_dose}$\rangle$ and a ground-truth tuple $\langle$``Cobalamin'', ``2.5'', ``$\mu$g'', \texttt{per\_daily\_dose}$\rangle$ would be rejected by the fast tier on the nutrient field --- the strings do not match after normalisation --- but accepted by the LLM tier on the grounds that ``Cobalamin'' and ``Vitamin B12'' name the same molecule.

Two design choices about the LLM tier deserve naming. First, the LLM is used only as an evaluator and never as an extractor. The pipeline does not consult an LLM during inference; the LLM only enters at scoring time, and only on pairs the deterministic tier has already rejected. This isolation is what allows the cost of the LLM tier to be bounded and reported. Second, the LLM judge is a relatively small open-weight model running locally rather than a frontier API model. The choice favours reproducibility, cost, and data control over absolute judging accuracy, and it follows the convention in zero-shot evaluation that the judge should be at least one capability tier below the strongest available model so that the evaluation does not become an artefact of a single proprietary system.

\subsection{Per-Image Variance and Diagnostic Logging}\label{sec:bg-evaluation-diagnostics}

Aggregate F1 across the dataset is a single number, and a single number can hide the structure of the pipeline's failures. An aggregate 4F~F1 of 41.5 percent on the supplement-label evaluation set could in principle reflect any number of underlying distributions, and on this dataset it reflects a wide spread: a substantial set of images on which the pipeline reaches 80 percent or higher, a larger majority on which it lands in the middle of the range, and a tail of images on which it produces nothing at all because the OCR is unrecoverable, the layout is unusually-arranged, or the structured path and the paragraph fallback both fail. The aggregate alone hides this spread. The thesis therefore reports per-image breakdowns alongside aggregate metrics, and the experimental chapter treats the per-image distribution as a first-class object of analysis --- examining where the pipeline does well, where it fails, and why.

Four diagnostic CSV files are produced for every experiment. \texttt{evaluation\_summary.csv} records the aggregate metrics --- precision, recall, 3F~F1, 4F~F1, and \texttt{context\_cost\_pp} --- together with the experiment name, the timestamp, and the configuration. \texttt{per\_image\_results.csv} records the same metrics broken out one row per image, allowing per-image comparison across experiments and ranking of best- and worst-performing images. \texttt{tuples.csv} records every predicted tuple in long format, allowing inspection of the raw output without re-running the pipeline. \texttt{pair\_details.csv} records every (predicted, ground-truth) pair the matching algorithm considered, including which fields agreed under the predicates of \S\ref{sec:bg-evaluation-tuplef1}, which is the file from which a per-field error analysis can be reconstructed.

Together, the four files are the audit trail that the modular-and-interpretable commitment of \S\ref{sec:bg-problem} promised. From an aggregate F1 number a reader can drill into the per-image distribution; from a regressed image they can drill into the predicted tuples; from a wrong tuple they can drill into the pair-level matching; and from there they can trace the failure backwards through the pipeline to the OCR token that produced it. Every claim in the experimental chapter is supported by at least one of these files, and the diagnostic logs are committed alongside the experimental results so that a reader who disputes a claim can reproduce the supporting evidence without re-running the pipeline.

% =========================================================================
\section{Modular Pipeline Design Philosophy}\label{sec:bg-philosophy}
% =========================================================================

The technical content of Chapter~\ref{ch:background} has now been laid out across the preceding sections. Each one introduced a specific component or concept the pipeline depends on: OCR and post-OCR repair in \S\ref{sec:bg-ocr}, geometric primitives in \S\ref{sec:bg-layout}, classifier families in \S\ref{sec:bg-classification}, graph reasoning in \S\ref{sec:bg-graphs}, embedding-based prototype matching in \S\ref{sec:bg-embeddings}, and the evaluation regime in \S\ref{sec:bg-evaluation}, all anchored in the commitments stated at the outset in \S\ref{sec:bg-problem}. This closing section steps back from individual components and articulates the architectural philosophy that runs through the chapter as a whole. The philosophy is not a new technique introduced here; it is the connective tissue between everything Chapter~\ref{ch:background} has already covered, named explicitly so that the reader closes the chapter with a coherent mental model of how the pipeline is organised and why.

The first principle is \textbf{stage isolation through typed contracts}. The pipeline is organised as a sequence of stages --- OCR and repair, classification and enrichment, graph construction, association, evaluation --- and each stage has a fixed input schema and a fixed output schema that constitute its contract with the surrounding pipeline. No stage reaches into another stage's internal state; no stage depends on the implementation choices of any other. The practical consequence is that single-component ablation experiments yield interpretable evidence: a stage can be replaced with a different implementation, the rest of the pipeline left untouched, and any change in pipeline output attributed unambiguously to the replaced stage. The alternative to typed contracts --- a tightly coupled pipeline where stages share state --- would compromise the interpretability of ablation results, because every replacement would risk silent ripple effects elsewhere in the pipeline.

The second principle is \textbf{diagnostic logging at every stage}. Beyond the data each stage emits to the next, each stage also emits a structured record of the decisions it made internally: which correction rules fired, which classifier rules were tried, which edges the graph constructor pruned, which pairs the matcher committed and at what score. The four CSV files introduced in \S\ref{sec:bg-evaluation-diagnostics} --- \texttt{evaluation\_summary.csv}, \texttt{per\_image\_results.csv}, \texttt{tuples.csv}, \texttt{pair\_details.csv} --- represent one externalised view of this logging, and together with the stage-level logs they aggregate, they form an audit trail in which every output tuple can be traced backwards to the OCR tokens that produced it, through the specific decisions that committed each link in the chain. The audit trail is what allows a regression on a single image to be localised to a specific stage, a specific token, and a specific decision, rather than being attributable only to ``something changed somewhere''.

The third principle is \textbf{deterministic implementations evaluated against learned alternatives}. Three of the chapter's sections name a deterministic-versus-learned trade-off explicitly: \S\ref{sec:bg-class-zeroshot} (rule-based versus learned NER versus embedding-based prototype matching for classification), \S\ref{sec:bg-graphs-detlearn} (deterministic versus learned graph reasoning), and \S\ref{sec:bg-embeddings} (the BGE-M3-driven embedding classifier as a learned alternative to rule-based nutrient detection). At each of these decision points, the chapter has presented the deterministic and learned options as parallel candidates and identified the empirical comparison as a question the experimental chapter examines. The thesis does not claim in this chapter that one regime is preferable to the other in general; the choice depends on the specific domain, the available data, and the constraints the application operates under. What the chapter does establish is that the trade-off is real, that it appears at multiple stages of a complex pipeline, and that the design space at each stage admits more than one defensible choice.

Together, these three principles --- stage isolation, diagnostic logging, and the deterministic-versus-learned framing --- are what the pipeline of this thesis is designed to embody, and what the methodology and experimental chapters of the thesis will draw on as they describe the pipeline's specific implementations and report the empirical results those implementations produced.